Distinct goal location beta frequency dynamics in hippocampus and prefrontal cortex across learning

Version 1

Abstract

Neural activity at goal locations contributes to learning by providing feedback on the success of preceding actions. This period engages neocortical and hippocampal networks, which serve distinct functions in processing reward and forming associations with experience that lead to reward. A neocortical network signature for reward feedback processing is beta oscillations (15-30Hz). Beta oscillations are thought to coordinate distributed neural processes across brain regions. However, it is unknown whether beta oscillations coordinate hippocampal-neocortical networks during the goal period, or how their dynamics relate to learning. Here, we show that beta oscillations occur in both hippocampal CA1 and the prefrontal cortex (PFC) when rats reach goal locations in spatial navigation tasks. Despite the presence of beta oscillations in both regions after goal entry, beta activity in each region differed in spectral and temporal properties. These differences suggest that the hippocampus and PFC are not strongly coupled at the beta frequency. We found the strengths of PFC and CA1 beta oscillations across learning were inversely related: PFC beta activity increasing in strength and CA1 beta activity decreasing in strength. Beta burst properties in PFC also had an inverse relationship to those of hippocampal sharp wave-ripples (SWRs), a prominent hippocampal process required for learning. We found a subset of PFC neurons modulated by both beta and hippocampal SWRs, which had distinct task-related firing patterns. Our results suggest that during outcome processing at goal locations, the neocortex and hippocampus are independently modulated by beta oscillations before becoming coordinated for memory-related processes during SWRs.

Introduction

Learning leads to changes in distributed neural networks across the brain. During learning, diverse neural processes across brain networks need to be coordinated, which is hypothesized to rely on neural oscillations that rhythmically couple activity in connected brain regions. This coupling between groups of brain regions is thought to be dynamic, creating functionally interacting subnetworks to support specific neural processes when required. A neural oscillation that is hypothesized to support these functions is beta (15-30Hz) (Bibbig et al., 2002; Gourévitch et al., 2010; Igarashi et al., 2014; Jayachandran et al., 2023; Kopell et al., 2011; Lundqvist et al., 2024; Miles et al., 2023; Spitzer & Haegens, 2017; Symanski et al., 2022). Beta oscillations occur in many brain regions, including sensory (Gervais et al., 2007; Kay et al., 1996; Martin et al., 2004), neocortical (Baker et al., 1997; Buschman et al., 2012; Cohen et al., 2007; Cohen et al., 2015; Dubey et al., 2023; HajiHosseini & Holroyd, 2015; HajiHosseini et al., 2020; Iturra-Mena et al., 2023; Kawasaki & Yamaguchi, 2013; Koloski, Hulyalkar, et al., 2024; MacKay & Mendonça, 1995; Pesaran et al., 2008; Voloh et al., 2020; Walsh et al., 2025; Xiao et al., 2024), subcortical (DeCoteau et al., 2007; Feingold et al., 2015; Howe et al., 2011; Lansink et al., 2009; Leventhal et al., 2012; Quinn et al., 2010; Sturman & Moghaddam, 2012) and limbic regions (Amaya et al., 2024; Berke et al., 2008; Fourcaud-Trocmé et al., 2019; França et al., 2014; França et al., 2021; Igarashi et al., 2014; Jackson et al., 2024; Jayachandran et al., 2023; Kay & Freeman, 1998; Rangel et al., 2015; Rangel et al., 2016; Samson et al., 2017; Symanski et al., 2022). Beta oscillations are associated with a wide range of cognitive functions, including sensory perception (Kay, 2014), reward processing (Cohen et al., 2007), top-down control (Engel & Fries, 2010) and novelty detection (Marco-Pallarés et al., 2015). Despite their widespread occurrence and association with diverse cognitive functions, a less-understood aspect of beta oscillations is their contribution to modulating or coordinating brain networks during learning, when rules are not fully understood and performance is improving. Most studies investigating beta dynamics across the brain have primarily focused on performance after initial learning, but three lines of work in sensory-limbic, striatal, and amygdala networks point to beta playing a role in supporting learning. Sensory-association learning studies suggest that beta activity coordinates sensory-limbic networks to enable the formation of associations between the sensory cue and its behavioral relevance (Fourcaud-Trocmé et al., 2019; Martin et al., 2007; Martin et al., 2004), beta oscillations dynamics in the ventral striatum change as habits are formed (Howe et al., 2011), and beta oscillations in the basolateral amygdala can drive reward reference (Amaya et al., 2024). However, the patterns of learning-related changes in beta oscillations across brain networks remain unclear, since the dynamics of beta during learning outside of olfactory and striatal networks are not known. Here, we ask how beta dynamics contribute to spatial learning in hippocampal-neocortical networks, which are brain regions known to play central roles in supporting memory processing and guiding behavior.

To understand how beta oscillations in hippocampal-neocortical networks change with learning, we focus on an important period for driving learning: when the reward outcome is revealed. We refer to this period as the “goal” phase of a task. Beta oscillations have been observed during equivalent task periods in both human and non-human animal studies. Neocortical beta oscillations during this period have been associated with processing outcomes, where feedback from the presence or absence of the reward can inform future actions and drive learning (Cohen et al., 2011). Goal-related beta oscillations have been observed in frontal cortical regions (Cohen et al., 2007; Kawasaki & Yamaguchi, 2013), including the anterior cingulate cortex (ACC) (Iturra-Mena et al., 2023; Xiao et al., 2024), the prefrontal cortex (PFC) (HajiHosseini & Holroyd, 2015; HajiHosseini et al., 2020; Voloh et al., 2020) and the orbitofrontal cortex (OFC) (Koloski, Hulyalkar, et al., 2024; Koloski, O’Hearn, et al., 2024), in primate and rodent species. Beta oscillations in these frontal cortical regions have been hypothesized to provide top-down cortical control of connected networks (Babapoor-Farrokhran et al., 2017; Benchenane et al., 2011; Bressler & Richter, 2015; Brovelli et al., 2004; Buschman et al., 2012; Dubey et al., 2023; Engel & Fries, 2010; Pesaran et al., 2008).

Despite frontal cortical beta activity being implicated in outcome feedback processing, whether frontal cortical and hippocampal networks are coordinated in the beta frequency range during the goal period remains unknown. Beta frequency coordination of hippocampal-neocortical networks has been reported in olfactory sensory decision-making tasks during the cue presentation period (Igarashi et al., 2014; Jayachandran et al., 2023; Martin et al., 2007; Symanski et al., 2022), and is hypothesized to support the use of memory-based sensory associations for decision making (Spitzer & Haegens, 2017). However, goal-related beta frequency oscillations have not been reported in the hippocampus. Beta oscillations in the hippocampus have so far been observed during the exploration of novel environments (Berke et al., 2004), interaction with novel objects (França et al., 2014; França et al., 2021; Rangel et al., 2015), approaches to goals (Iwasaki et al., 2021; Lansink et al., 2009) as well as sensory stimulus sampling (Fourcaud-Trocmé et al., 2019; Gourévitch et al., 2010; Jayachandran et al., 2023; Kay & Freeman, 1998; Kay et al., 1996; Martin et al., 2007; Rangel et al., 2016; Symanski et al., 2022). Thus, it is unknown if and how hippocampal-neocortical networks are coordinated at the beta frequency during the goal phase of a task, and how this coordination relates to learning.

To understand the relationship between goal-related beta frequency activity in hippocampal-neocortical networks and learning, we recorded local field potentials (LFPs) and single units from the hippocampus (dorsal CA1) and the prefrontal cortex (PFC) in rats learning spatial navigation tasks over five days. We observed beta oscillations in both regions at the goal locations of the tasks. However, we were surprised to find that beta oscillations in the PFC and CA1 had distinct spectral and temporal properties. The mismatch in frequency range and timing is consistent with each region engaging in distinct beta-activity modulation, likely with other brain regions, rather than being coupled with each other by beta. Over days of training, beta oscillations in the PFC and CA1 had distinct correlations with performance, where PFC beta increased in strength and CA1 beta decreased in strength. We also found an inverse relationship between the learning-related properties of PFC beta and a prominent hippocampal marker of learning, the sharp wave-ripple (SWR). Lastly, we found distinct task correlates in a subpopulation of PFC cells that showed modulation by both beta oscillations and hippocampal SWRs.

In contrast to previous results demonstrating coordinated hippocampal-neocortical beta modulation during sensory decision-making (Gourévitch et al., 2010; Jayachandran et al., 2023; Kay & Freeman, 1998; Kay et al., 1996; Martin et al., 2007; Rangel et al., 2016; Symanski et al., 2022), we show that beta oscillations in these networks are likely modulating distinct network processes during the goal phase of a task. Further, beta dynamics in the PFC and CA1 undergo opposite changes across learning. After goal entry, the neocortex and hippocampus are uncoupled at the beta frequency but later coordinate during hippocampal SWRs. Our results support the hypothesis that at the goal, the neocortex is initially engaged in processing outcome-related information before coordinating with hippocampal networks for memory-related processes.

Results

Rats (n=6) were trained on two spatial alternation tasks, interleaved across sessions, for five days. In each task, reward was delivered when the three goal locations were visited in an alternating sequence (Fig. 1A). We first asked whether beta oscillations were present in both CA1 and PFC (Fig. 1-1). We saw power in the beta-frequency range (15-30Hz) in both PFC and CA1 after the animal reached the goal locations (Fig. 1B). This was confirmed by the power spectral density (PSD) for periods when the animal was at the goal location (Fig. 1C). The PSD from the PFC showed a peak at approximately 20Hz. Although the PSD in CA1 did not show a peak in the beta frequency range like the one found in PFC, the continuous wavelet transform from single trials showed beta oscillations occurred in bursts (Fig. 1B), which is consistent with beta in other neocortical regions (Sherman et al., 2016; Shin et al., 2017). To characterize the properties of beta-frequency bursts in PFC and CA1, we extracted burst times from the continuous wavelet transform (Fig. 1-2). We found the spectral properties of PFC and CA1 bursts differed significantly. PFC bursts were tightly distributed in frequency (median peak frequency: 19 Hz), whereas the peak frequencies of CA1 bursts were distributed across a wider frequency range (median peak frequency: 22 Hz) (Fig. 1D-E). This explains the peak in the PFC PSD and the lack of such a peak in the CA1 PSD. The burst amplitude profiles also differed between the two regions. In PFC, lower-frequency bursts tended to be higher in amplitude, whereas CA1 bursts were more uniform in size (Fig. 1-3A). Burst durations were similar across the two regions (Fig. 1-3B).

Distinct PFC and CA1 beta frequency activity after the reward location entry during spatial navigation tasks.

a. Spatial learning tasks on Y and F shaped mazes with spatial alternation rules. Animals were trained on the mazes across interleaved sessions. Reward locations (1, 2 and 3) are indicated. The order of visits required to ensure reward on every trial is shown above the maze schematics. The daily training schedule shows the duration, order and separation between sessions. b. Example PFC (orange) and CA1 (blue) LFP traces aligned to goal location entry (black vertical line). The beta frequency-filtered LFP is shown below the raw LFP. The corresponding continuous wavelet transform spectrogram is included. Beta bursts are indicated by horizontal lines. Dotted vertical lines mark 1 s intervals. c. Power spectral density for time intervals when the animal is at the reward locations. Left panel shows the PSD, with the aperiodic component shown in gray. Right panel show PSD with aperiodic component subtracted. The peak marked by * in the CA1 PSD at ∼15Hz corresponds to the first harmonic of theta (Scheffer-Teixeira & Tort, 2016). d. Frequency profile for beta bursts grouped by peak power frequency for PFC (orange) and CA1 (blue). Top: the vertical histogram of the mean frequency profiles for bursts at each peak power frequency. Bottom: histograms with the count distribution for bursts within 15-30Hz, grouped by peak power frequency. Shading corresponds to the probability value at each frequency. Bursts occurring within 5s after reward location entry are included. e. Cumulative distribution of peak burst frequencies for PFC (orange) and CA1 (blue). Kolmogorov-Smirnov test, p=2.9×10−103. f. Distribution of burst times relative to goal location entry grouped by peak frequency for PFC (orange) and CA1 (blue). g. Cumulative distribution of burst occurrence relative to goal location entry for PFC (orange) and CA1 (blue). Kolmogorov-Smirnov test, p=1.1×10−4.

Location of recording tetrodes.

a. Representative histology showing location of tetrodes in PFC. b. Representative histology showing location of tetrodes in CA1.

Burst detection.

a. Two example bursts with corresponding LFP and continuous wavelet transform from PFC (1.5 s each, top row). The burst is marked by the black horizontal line. The corresponding continuous wavelet transform is shown in the middle row. The white cross marks the frequency and time corresponding to the maximum power in the burst interval. The mean power in the beta band (15-30Hz) from the continuous wavelet transform is shown by the histogram in the bottom row. The solid black line corresponds to 1 standard deviation (z) above the mean (1z), and the dotted black line corresponds to 0.5 standard deviation above the mean (0.5z). To detect bursts, we identify intervals during which the beta-band power exceeds 0.5 standard deviation above the mean. Intervals are included if the maximum power in the interval exceeds 1 standard deviation above the mean, and the duration of the interval is greater than 100ms. Intervals separated by less than 100ms were merged. The peak frequency for each burst is the frequency corresponding to the maximum power within the burst interval.

Power and duration for beta bursts by frequency.

a. Beta burst power by peak burst frequency. b. Beta bursts duration by peak burst frequency.

Beta bursts occur independently in both regions.

a. Beta frequency bursts (shaded) detected across multiple electrodes in PFC (orange) and CA1 (blue) relative to reward location entry (black vertical line). Three example trials are shown. b. Coherence (mean±SEM) within PFC (orange), between PFC and CA1 (green), or within CA1 (blue) for periods at the goal location. c. Mean coherence (15-30Hz) across the groups. Pairwise rank sum test with Benjamini-Hochberg false discovery rate correction: p<0.001 for all pairs. d. Cross-correlation (mean±SD) of burst peak times for bursts detected on tetrodes with PFC (orange), across PFC-CA1 (green) and within CA1 (blue). e. Proportion of the cross-correlation density within ±50ms lag for PFC (orange), across PFC-CA1 (green) and within CA1 (blue). Pairwise rank sum test with Benjamini-Hochberg false discovery rate correction: PFC versus PFC-CA1 (p=0.003**), PFC versus CA1 (p=0.24n.s.) and PFC-CA1 versus CA1 (p=0.003**). f. Frequency distribution of bursts by peak frequency for independent and coincident bursts in each region. g. Burst peak frequency distribution for coincident and independent bursts. Cumulative distribution burst occurrence relative to goal location entry. Kolmogorov-Smirnov test: PFC (p=0.96n.s.) and CA1 (p=0.002**).

Beta burst aligned power and coherence

a. Multi-taper power spectrum within a 1 s window centered on the peaks of independent or coincident beta bursts (columns) in the PFC (top row) and CA1 (bottom row). The heatmap shows the mean across all bursts in each category. b. Power (18-25Hz) within ±100 ms of the peak. PFC (orange, top) and CA1 (blue, bottom). Pairwise rank-sum test with Benjamini-Hochberg false discovery rate correction: p<0.001*** and p<0.05* for all significant comparisons. c. Change in power within ±100 ms of the peak relative to baseline (125 ms from - 0.5 s). PFC (orange, top) and CA1 (blue, bottom). Wilcoxon signed-rank test with Benjamini-Hochberg false discovery rate correction: p<0.001***. d. Multi-taper coherence in PFC (upper row) and CA1 (lower row) within a 1s window centered on the peaks of independent or coincident beta bursts. The heatmap shows the mean across all bursts in each category. e. Coherence (18-25Hz) within ±100 ms of the peak in PFC (upper row) and CA1 (lower row). Pairwise rank sum test with Benjamini-Hochberg false discovery rate correction: p<0.001*** and p<0.05* for all significant comparisons. f. Change in coherence within ±100ms of the peak relative to baseline (125ms from -0.5s). PFC (orange, top) and CA1 (blue, bottom). Wilcoxon signed-rank test with Benjamini-Hochberg false discovery rate correction: p<0.001***. g. PFC-CA1 coherence within a 1 s window centered on the peaks of independent or coincident beta bursts. The heatmap shows the mean across all bursts in each category. h. PFC-CA1 coherence between (18-25Hz) within ±100 ms of the peak in PFC (top row) and CA1 (bottom row). Pairwise rank sum test with Benjamini-Hochberg false discovery rate correction: p<0.001*** and p<0.01** for all significant comparisons. i. Change in PFC-CA1 coherence within ±100 ms of the peak relative to baseline (125 ms from - 0.5 s). PFC (orange, top) and CA1 (blue, bottom). Wilcoxon signed-rank test with Benjamini-Hochberg false discovery rate correction: p<0.001*** and p<0.01**.

Distinct changes in PFC and CA1 burst properties across days of learning.

a. Performance in each maze task by day. Linear model with fixed (day) and random (animal) effects. Y maze (βday=0.073, SE=0.001, 95% CI=(0.071, 0.075), z=71.9 and p<0.001***) and F maze (βday=0.04, SE=0.001, 95% CI=(0.038, 0.042), z=42.4 and p<0.001***). b. Burst power in each maze task by day. Linear model with fixed (day) and random (animal) effects. Y maze PFC (βday=347.7, SE=38.2, 95% CI=(272.9, -422.5), z=9.1 and p<0.001***) and F maze PFC (βday=300.3, SE=49.4, 95% CI=(203.5, 397.1), z=6.1 and p<0.001***). Y maze CA1 (βday=-1133.2, SE=121.5, 95% CI=(-1371.3, -895.0), z=-9.3 and p<0.001***) and F maze CA1 (βday=-711.7, SE=158.6, 95% CI=(-1022.6, -400.9), z=4.5 and p<0.001***). c. Burst peak frequency variance in each maze task by day. Linear model with fixed (day) and random (animal) effects. Y maze PFC (βday=-1.1, SE=0.3, 95% CI=(-1.7, -0.48), z=-3.6 and p<0.001***), F maze PFC (βday=-0.37 SE=0.3, 95% CI=(-.94, .22), z=6.1 and p=0.22n.s.), Y maze CA1 (βday=-0.89, SE=0.30, 95% CI=(-1.5, -0.31), z=-3.0 and p=0.003**) and F maze CA1 (βday=-0.38, SE=0.32, 95% CI=(-1.0, 0.25), z=-1.2 and p=0.24n.s.) .

The timing of bursts relative to goal entry also differed between PFC and CA1 (Fig. 1F-G). Bursts tended to start approximately 400ms after reward location entry, but CA1 bursts were more concentrated within the first two seconds compared with PFC bursts, which showed a sustained pattern. The heatmap shows a clear difference in the burst frequency distribution between the two regions and their distribution over time. Although beta is generally defined as oscillations in the 15-30Hz range, it has been separated into a lower-frequency beta1 (∼15Hz) and a higher-frequency beta2 (20-30Hz), based on work in cortical regions (Baker et al., 1997; MacKay & Mendonça, 1995; Roopun et al., 2008; Roopun et al., 2006). However, the frequency of PFC beta bursts described here is higher than beta1 and lower than beta2. The frequency of CA1 beta bursts is consistent with beta2 previously described in the hippocampus (Berke et al., 2008; Igarashi et al., 2014; Iwasaki et al., 2021). These results indicate that PFC and CA1 beta bursts at the goal have distinct spectral and temporal properties.

Beta oscillations are hypothesized to coordinate neural activity across brain regions (Spitzer & Haegens, 2017) to support cognitive processes including sensory-guided decisions and reinstatement of context-dependent memory patterns (Igarashi et al., 2014; Jayachandran et al., 2023; Kay & Freeman, 1998; Symanski et al., 2022). We therefore asked whether beta oscillations can coordinate PFC and CA1 activity at the goal. If so, we would expect coherence in the beta-frequency range between PFC and CA1, and similar burst timing in both regions. Visual inspection of LFP signals from PFC and CA1 tetrode showed that some beta bursts were coincident, while others appeared independent in time (Fig. 2A). To quantify beta-frequency coordination between regions, we first calculated the LFP coherence between tetrodes within each region as well as across regions, for periods when the animal was at the goal. We found higher coherence within PFC or CA1 compared to coherence between PFC and CA1 (Fig. 2B-C). The PFC coherence within the beta range showed a peak centered at approximately 20Hz, whereas the CA1 coherence was similar across the entire beta-frequency range. The across-region coherence was significantly weaker but still showed a small peak at ∼20Hz. We then analyzed burst-wise timing and found beta bursts were more coincident within each region compared with across regions (Fig. 2D-E). The cross-correlation of burst peak times showed higher values around 0s lag within each brain region compared with the cross-correlation between PFC and CA1.

Our results so far suggest beta-frequency oscillations are distinct across PFC and CA1 based on their frequency distributions and low burst coincidence. Despite these differences, it is possible that a subset of coincident PFC and CA1 bursts could coordinate activity across the two regions. This hypothesis predicts that coincident bursts have similar frequency distributions and are more coherent between the two regions. We therefore classified each burst as “coincident” across regions, if it overlapped in time with a burst in the other brain region. Bursts that did not overlap were classified as “independent” across the regions. Based on this method, we asked whether coincident bursts were more closely matched in their frequency distribution across the regions compared with independent bursts. However, this was not the case, as coincident bursts in CA1 were higher in peak frequency compared with coincident bursts in PFC (Fig. 2F-G). The difference in frequency distribution for the coincident bursts provides evidence against a common beta-frequency range coordinating PFC and CA1, since there is a significant frequency mismatch. We confirmed our classification of independent and coincident bursts using burst-aligned power and coherence analysis (Fig. 2-1). Independent bursts in one region were not accompanied by significant increases in beta power in the other region (Fig. 2-1A-B). Coincident bursts, by definition, showed the expected increase in power in both regions. We saw a similar pattern for coherence (Fig. 2-1D-E). Coherence within each region increased during independent bursts but not in the other region. Coherence in both regions increased during coincident bursts. Coherence across regions also increased for coincident bursts more than independent bursts (Fig. 2-1G-H), however, the coherence value across brain regions was lower in comparison with coherence within each region (Fig. 2-1E vs. Fig. 2-1H).

Having established the properties of beta oscillations in PFC and CA1, we next asked how goal-related hippocampal-neocortical beta oscillations changed across task learning. We therefore determined the relationship between beta properties in PFC and CA1 with either days of training or performance in each task. Animals performed better in the Y alternation compared with the F alternation over five days (Fig. 3A). By day five, performance reached greater than 80% on the Y compared with approximately 60% on the F maze tasks, indicating the F maze task was more challenging. This difference in performance allowed us to examine the effects of both training time and performance improvement on beta dynamics. In both tasks, PFC beta-burst power increased over the five days, whereas CA1 beta-burst power decreased (Fig. 3B). Interestingly, beta bursts became more consistent in their peak frequency across days, but only when performance improved. The variance in PFC and CA1 burst frequency decreased over days of training for the Y but not the F maze (Fig. 3C). We repeated the same set of analyses by partitioning the data based on performance quintiles (Fig. 4A), which allowed us to directly relate beta activity properties with performance. We found similar relationships between burst power (Fig. 4B) and frequency variance (Fig. 4C).

Distinct changes in PFC and CA1 burst properties with performance.

a. Performance in each maze task by performance quintile. Linear model with fixed (performance quintile) and random (animal) effects. Y maze (βday=0.084, SE=0.001, 95% CI=(0.083, 0.086), z=103.3 and p<0.001***) and F maze (βday=0069, SE=0.001, 95% CI=(0.068, 0.070), z=136.9 and p<0.001***). b. Burst power in each maze task by performance quintile. Linear model with fixed (performance quintile) and random (animal) effects. Y maze PFC (βperf.=484.4, SE=40.5, 95% CI=(405.1, 563.8), z=12.0 and p<0.001***) and F maze PFC (βperf.=242.6, SE=60.9, 95% CI=(123.2, 362.0), z=4.0 and p<0.001***). Y maze CA1 (βperf.=-1199.6, SE=127.2, 95% CI=(-1448.9, -950.3), z=-9.4 and p<0.001***) and F maze CA1 (βperf.=60.6, SE=198.3, 95% CI=(-328.1, 449.3), z=0.3 and p=0.76n.s.). c. Burst peak frequency variance in each maze task by performance quintile. Linear model with fixed (performance quintile) and random (animal) effects. Y maze PFC (βperf.=-0.80, SE=0.30, 95% CI=(-1.4, -0.22), z=--2.7 and p<0.007**) and F maze PFC (βperf.=-0.68, SE=0.54, 95% CI=(-1.74, 0.39), z=-1.2 and p=0.21n.s.). Y maze CA1 (βperf.=-1.2, SE=0.35, 95% CI=(-1.87, -0.50), z=-3.40 and p=0.001**.) and F maze CA1 (βperf.=-0.27, SE=0.67, 95% CI=(-1.57, 1.03), z=-0.41 and p=0.68n.s.).

We next found beta bursts occurred earlier after reaching the goal location over days of training (Fig. 5), pointing to temporal changes in beta-frequency modulated processes in PFC and CA1 with learning. We examined rewarded and unrewarded trials separately. During early learning days, bursts in PFC and CA1 on rewarded trials tended to occur several seconds after entry to the goal (Fig. 5A-E). During later days of training, the peak in the probability density shifted to earlier time points relative to goal entry. The same trend across days was found for unrewarded trials, but more strikingly, bursts on unrewarded trials occurred earlier than those on rewarded trials (Fig. 5C, F). These results demonstrate that beta frequency modulation in PFC and CA1 has distinct correlates with performance across the five days of training, suggesting changes to beta frequency activity reflect learning-related dynamics in PFC and CA1 networks.

Bursts occur sooner after goal entry over days of training.

a. Y maze burst distribution after goal entry for rewarded trials. PFC (orange) and CA1 (blue). Gray line indicates the smoothed histogram. Black plus marks the peak of the smoothed histogram. Linear model with fixed (day) and random (animal) effects on burst timing: PFC (βday=-0.127 SE=0.023, 95% CI=(-0.17, -0.081), z=-5.45 and p<0.001) and CA1 (βday=-0.11 SE=0.030, 95% CI=(-0.17, -0.048), z=-3.54 and p<0.001). b. Y maze burst distribution after goal entry for unrewarded trials. PFC (orange) and CA1 (blue). Linear model with fixed (day) and random (animal) effects on burst timing: PFC (βday=-0.14 SE=0.057, 95% CI=(-0.26, -0.031), z=-2.51 and p=0.012) and CA1 (βday=-0.23 SE=0.065, 95% CI=(-0.34, -0.099), z=-3.48 and p<0.001). c. Boxplot comparison for rewarded and unrewarded trials. Wilcoxon rank sum test: p=2.20×10−4 (PFC) and p=1.15×10−2 (CA1). d. F maze burst distribution after goal entry for rewarded trials. PFC (left) and CA1 (right). Gray line indicates the smoothed histogram. Black plus marks the peak of the smoothed histogram. Linear model with fixed (day) and random (animal) effects on burst timing: PFC (βday=-0.042 SE=0.037, 95% CI=(-0.12, 0.030), z=-1.15 and p=0.25) and CA1 (βday=-0.05 SE=0.047, 95% CI=(-0.18, 0.008), z=-1.78 and p=0.073). e. F maze burst distribution after goal entry for unrewarded trials. PFC (left) and CA1 (right). Linear model with fixed (day) and random (animal) effects on burst timing: PFC (βday=-0.082 SE=0.043, 95% CI=(-0.17, 0.003), z=-1.9 and p=0.057) and CA1 (βday=-0.056 SE=0.058, 95% CI=(-0.17, -0.057), z=-0.97 and p=0.33). f. Boxplot comparison for rewarded and unrewarded trials. Wilcoxon rank sum test: p=2.89×10−4 (PFC) and p=8.01×10−4 (CA1).

We next asked how neurons in PFC and CA1 are modulated by beta-frequency oscillations. We found that cells in PFC and CA1 were more strongly phased-locked to beta oscillations within their own region (local) rather than the beta oscillation in the other region (remote). Both PFC (20%) and CA1 (41%) populations showed significant phase locking to beta within each region (Fig. 6A-C). The PFC and CA1 cell populations showed different phase preferences to local beta, with PFC cells showing a bias toward spiking during the rising phases, whereas CA1 cells showed a slight bias for the falling phases (Fig.6D). Further, we found cells in each region showed higher phase locking to local beta compared with remote beta (Fig. 6E-F). This indicates that spiking in each region is modulated preferentially by local beta oscillations rather than remote beta oscillations. These results provide further evidence supporting that the PFC and CA1 are independently modulated at the beta frequency during the goal period.

PFC and CA1 cells show stronger spiking modulation to local beta.

a. Example PFC (orange) and CA1 (blue) spike phase locking histogram, sorted by increasing phase locking strength. The white line indicates the peak and trough of beta. The distributions were duplicated across two cycles of beta to improve visualization. The strength of the concentration parameter kappa (k) is indicated for each cell. b. Distribution of Rayleigh z for PFC (upper) and CA1 (lower). The proportion of cells showing significant phase locking are shown. c. Distribution of k from a von Mises distribution fit for significantly beta modulated cells (Rayleigh p <0.05 in panel b) in PFC (upper) and CA1 (lower). d. Distribution of phase preference for cells in PFC (upper) and CA1 (lower). Red dotted line shows the mean population phase preference. The gray line indicates the peak and trough of beta. Watson-Williams circular means test for PFC and CA1: p=1.01×10−10. e. Beta phase locking strength (Rayleigh z) to local or remote beta for cells in PFC (upper) and CA1 (lower). Wilcoxon signed-rank test: p=3.17×10−15 (PFC) and p=3.64×10−31 (CA1). f. Beta phase locking strength (kappa) to local or remote beta for cells in PFC (upper) and CA1 (lower). Wilcoxon signed-rank test: p=2.20×10−11 (PFC) and p=6.76×10−24 (CA1).

The inverse learning-related changes in PFC and CA1 beta oscillations point to complementary processes modulated by beta in each region. We next asked how changes in beta oscillations are related to another prominent neural process supporting learning: awake hippocampal SWRs. Hippocampal SWRs also occur at goal locations (Ambrose et al., 2016; Carey et al., 2019; Diba & Buzsaki, 2007; Foster & Wilson, 2006; Kleinman & Foster, 2025; Singer & Frank, 2009) and are hypothesized to coordinate distributed representations (Jadhav et al., 2016; Shin et al., 2019; Tang et al., 2021; Yu et al., 2017; Yu et al., 2018) to support the formation and use of memory during spatial learning (Cheng & Frank, 2008; Deceuninck & Kloosterman, 2024; Dupret et al., 2010; Gupta et al., 2010; Jadhav et al., 2012; Pfeiffer & Foster, 2013; Singer et al., 2013; Singer & Frank, 2009). Since both beta and SWRs occur at goal locations, this led us to ask how beta-frequency dynamics within PFC and CA1 networks are related to hippocampal SWR dynamics. We found several features of beta-frequency oscillations that show the opposite pattern of change compared with SWRs. Consistent with previous findings, we found SWRs tended to occur several seconds after the animal reached the goal (Fig. 7A, B) (Ambrose et al., 2016; Carey et al., 2019; Diba & Buzsaki, 2007; Foster & Wilson, 2006; Kleinman & Foster, 2025; Singer & Frank, 2009) and the amplitude of SWRs decreased across days (Fig. 7C) (Cheng & Frank, 2008). In contrast to the delayed onset of hippocampal SWRs, beta bursts in PFC and CA1 occurred closer to the time of goal entry (Fig. 7A, D); thus, on average beta bursts in each region occurred earlier than SWRs. We confirmed this relationship occurred on a single-trial timescale by computing the cross-correlation between the timing of beta bursts in each region and SWRs (Fig. 7E). Further, the amplitude of PFC beta-frequency bursts increased across days of training (Fig. 3B) in contrast to the decrease in SWR amplitude (Fig. 7C). These differences in the properties of beta-frequency activity and SWRs indicate that these are two neural signatures for complementary learning-related processes during learning.

Beta bursts occur earlier than hippocampal SWRs at goal locations and have inverse learning-related changes in amplitude compared with PFC beta bursts.

a. PFC (orange) and CA1 (blue) LFP relative to goal entry (black vertical line). Beta frequency filtered signal is shown below the LFP. Ripple frequency (150-250Hz) filtered CA1 signal is shown in black. b. Goal entry aligned histogram for SWRs. c. SWR size across days of training. Linear model with fixed (day) and random (animal) effects: βday=-0.15 SE=0.01, 95% CI=(-0.16, -0.13), z=-15.3 and p<0.001. d. Goal entry aligned histogram for PFC beta (orange) or CA1 beta bursts (blue). e. Cross-correlation between PFC (left) or CA1 (right) beta bursts and SWRs. Wilcoxon signed-rank test p=3.12×10−114 (PFC) and p=5.64×10−78 (CA1.)

Since our results suggest that beta frequency activity and SWRs correspond to distinct processes at the goal, do they modulate distinct populations of PFC and CA1 cells? In PFC and CA1, we found cells modulated by both SWRs and beta (Fig. 8A, B). In PFC, 20% of cells showed modulation to beta and 43% showed modulation to SWRs. The subset of PFC cells that are modulated by both SWR and beta (11%) is greater than the expected proportion (8.6%) under the assumption that SWR and beta can modulate the population independently (Fisher exact test, p=0.021), although the size of the difference is small. We found PFC cells modulated by both beta and SWRs have distinct firing properties (Fig. 8C-D). For these cells, we found a positive relationship between the spike-phase preference and the direction of SWR modulation (Fig. 8C left). PFC cells with spiking preference for the falling phases of the beta cycle tended to show spiking inhibition during SWRs (negative SWR modulation index) (Fig. 8A, lower row) whereas cells with spiking preference for the rising phases of beta tended to be excited during SWRs (positive SWR modulation index) (Fig. 8A upper row). This relationship was not observed for PFC cells that are modulated by beta but not modulated by SWRs (Fig. 8C, right, Fig. 8-1A). Further, we found a negative relationship between spike-phase preference and the cell’s firing pattern relative to goal entry. PFC cells with spiking preference for the falling phases of beta tended to spike later after goal entry whereas cells with phase preference to the rising phase of beta were more active sooner after goal entry (Fig. 8D, left). This again was not observed for PFC cells that were modulated by beta but not SWRs (Fig. 8D, right, Fig. 8-1B, Fig. 8A, compare top and bottom rows). We repeated these analyses for the CA1 population but did not find any statistically significant relationships between beta phase preference and SWR modulation (Fig. 8E-H). In CA1, ∼40% of cells showed modulation to beta and 76% of cells showed modulation to SWRs. The subset of CA1 cells that are modulated by both SWR, and beta (31%) was similar to the expected proportion (30%) under the assumption of independence (Fisher exact test, p=0.983). Our results show that beta oscillations modulate a subpopulation of PFC neurons active at the goal, and some of these cells are coordinated with hippocampal SWRs. The PFC subset may not be coordinated with the hippocampus at the beta frequency, but they can still be linked with relevant hippocampal representations during SWRs. Together, this subset of PFC cells could form a distinct cell ensemble across hippocampal-cortical networks relevant for learning from goal outcome.

SWR and beta modulation in PFC and CA1.

a. Four example PFC cells showing SWR aligned spiking raster and histogram, spike beta-phase locking histogram and goal entry aligned firing histogram. Top two cells show spiking excitation around SWRs, and bottom two cells show spiking inhibition around SWRs. In the spike beta-phase locking histogram, the distribution is duplicated to improve visualization. The peaks and troughs for the beta cycle is shown in white. For the goal entry aligned firing histogram, the median firing density is shown by the vertical black line. This corresponds to 50% of the firing density relative to goal entry. b. PFC beta modulation strength (kappa) versus SWR modulation index. The four combinations of SWR (S) and beta (β) modulation (+: modulated and -: not modulated) and their corresponding proportions are indicated. c. PFC SWR modulation index versus beta phase locking preference for beta and SWR modulated cells (β+S+) (left) and SWR modulated but not beta modulated cells (β+S-) (right). The regression line is show in cyan. The gray shading indicates the peak and trough of beta. Two cycles are repeated for clarity. d. PFC median firing density from goal entry versus beta phase locking preference for beta and SWR modulated cells (β+S+) (left) and SWR modulated but not beta modulated cells (β+S-) (right). Median firing density is the time from goal entry corresponding 50% of the firing density within 5 seconds. e. Four example CA1 cells showing SWR aligned spiking raster and histogram, spike beta-phase locking histogram and goal entry aligned firing histogram. f. CA1 beta modulation strength (kappa) versus SWR modulation index. The four combinations of SWR and beta (β) modulation (+: modulated and -: not modulated) and their corresponding proportions are indicated. g. CA1 SWR modulation index versus beta phase locking preference for beta and SWR modulated cells (β+S+) (left) and SWR modulated but not beta modulated cells (β+S-) (right). The regression line is show in red. The gray shading indicates the peak and trough of beta. Two cycles are repeated for clarity. h. CA1 median firing density from goal entry versus beta phase locking preference for beta and SWR modulated cells (β+S+) (left) and SWR modulated but not beta modulated cells (β+S-) (right). Median firing density is the time from goal entry corresponding 50% of the firing density within 5 seconds.

Permutation test for the significance of the correlation coefficient.

a. Distribution of the permuted (gray) and observed (vertical line) for PFC SWR modulation index versus beta phase locking preference in Fig. 8C. We performed this additional permutation test to verify the observed value of the correlation coefficient is different to the chance distribution. b. Distribution of the permuted (gray) and observed (vertical line) for PFC median firing density from goal entry versus beta phase locking preference in Fig. 8D. c. Distribution of the permuted (gray) and observed (vertical line) for CA1 SWR modulation index versus beta phase locking preference in Fig. 8G. d. Distribution of the permuted (gray) and observed (vertical line) for CA1 median firing density from goal entry versus beta phase locking preference in Fig. 8H.

Discussion

Our results show beta-frequency oscillations occurred after entry to goal locations in both CA1 and PFC. However, several properties suggest that beta may not be strongly coordinating both regions at the goal. First, we found that beta bursts had distinct spectral and temporal properties in each region, where PFC beta bursts were lower in frequency than CA1 bursts. Second, PFC burst amplitude increased with training and performance, whereas CA1 burst amplitude decreased. The pattern of learning-related changes for PFC bursts was also opposite to those of SWRs. Lastly, beta activity and SWR modulated PFC cells identified a specific subpopulation with distinct task correlates, but not in CA1.

Our results suggest that the neocortex and hippocampus may participate in distinct functional networks at the goal, and beta oscillations with distinct frequencies may be supporting interactions between each region with their respective network partners. This contrasts with other task periods where the PFC and hippocampus are part of the same functional network modulated by beta, or other frequencies. For example, during spatial navigation, PFC cells show strong phase locking to hippocampal theta (Hyman et al., 2005; Jones & Wilson, 2005; Siapas et al., 2005; Yu & Frank, 2021; Zielinski et al., 2019), a frequency band that has been hypothesized to reflect coordination of distinct task-related representations across these structures during the execution of cognitive tasks. During olfactory sensory decision-making, beta-frequency coherence is observed across neocortical and hippocampal networks, which is thought to coordinate associative memory representations (Fourcaud-Trocmé et al., 2019; Igarashi et al., 2014; Jayachandran et al., 2023; Kay & Freeman, 1998; Martin et al., 2007; Sheriff et al., 2021; Symanski et al., 2022). During awake quiescence, activity in neocortical and hippocampal networks is transiently synchronized during SWRs (Jadhav et al., 2016; Remondes & Wilson, 2015; Shin et al., 2019; Tang et al., 2017; Yu et al., 2017; Yu et al., 2018). Coordination at these frequencies and task periods demonstrates hippocampal-neocortical interactions play a central role in supporting diverse cognitive functions. Our results show that the neocortex and hippocampus may be decoupled or only weakly coupled during the goal period at the beta frequency, and each region may be engaged with its own set of functionally connected networks (Cohen et al., 2011). This was surprising given the extensive coordination across these brain structures during other important task periods. We had expected that stronger coordination during this period would be important for learning-related changes across neocortical-hippocampal networks, similar to SWRs that occur at the goal.

Our results point to another insight on the complementary changes in hippocampal-neocortical beta dynamics. We saw an inverse pattern in the hippocampus and PFC during learning. These results indicate that beta-frequency activity in the neocortex and hippocampus, together with SWRs, are markers of complementary learning-induced changes in these networks, pointing to the contrasting contributions of the hippocampus and neocortex to learning. The hippocampus is hypothesized to play an important role in rapid changes during early learning, whereas the neocortex gradually changes on a longer timescale (Buzsáki, 1996; Mcclelland et al., 1995). In the hippocampus, SWRs are a neural signature of early learning, where larger amplitude events are observed during novel experience (Cheng & Frank, 2008). With learning, these events become less prominent. We found similar patterns for goal-related beta activity in the hippocampus, which decreases with performance. This mirrors novelty-related beta activity in the hippocampus that occurs during exploration of new environments or objects, which decreases with familiarity (Berke et al., 2008; França et al., 2014; França et al., 2021; Rangel et al., 2015). In contrast to hippocampal beta oscillations, we hypothesize that beta activity in the PFC reflects the reconfigured state for a given task as learning progresses, where beta amplitude and frequency uniformity increases with experience and performance. This increase parallels findings in the olfactory bulb and piriform cortex, where the strength of beta during odor sampling increases with performance proficiency in olfactory sensory decision-making tasks (Cohen et al., 2015; Fourcaud-Trocmé et al., 2019; Gervais et al., 2007; Martin et al., 2007; Martin et al., 2004), the ventral striatum (Howe et al., 2011), as well as the emergence of beta oscillations in the basolateral amygdala with the development of reward preference (Amaya et al., 2024).

The difference in the frequency distribution of hippocampal and PFC beta raises the question of whether cortical and hippocampal beta have similar origins. Given that beta oscillations have been found in connected cortical, subcortical, and hippocampal regions, do they arise from similar mechanisms? Experimental evidence and computational models indicate that beta oscillations could be generated locally within a network (Bitzenhofer et al., 2017; David et al., 2015; Fourcaud-Trocmé et al., 2011; Osinski & Kay, 2016; Roopun et al., 2008; Roopun et al., 2006; Sherman et al., 2016; Womelsdorf et al., 2014). Beta was originally partitioned into a lower-frequency ∼15Hz (beta1) band and a higher frequency 20-30Hz (beta2) band (Baker et al., 1997; MacKay & Mendonça, 1995; Roopun et al., 2008; Roopun et al., 2006). Our observations in CA1 are consistent with beta2, with beta bursts having peak frequencies primarily above 20Hz (Berke et al., 2008; Igarashi et al., 2014; Iwasaki et al., 2021). The PFC beta we observed centers around 19Hz with a narrow frequency distribution. This is closer to beta1 but is higher than the beta1 reported in various neocortical regions. Could the distinct beta frequency distribution in these regions point to distinct origins? PFC beta may be generated locally or from a brain region with a lower frequency beta generator. The wider distribution of CA1 beta burst peak frequencies suggests hippocampal beta may arise from multiple local or remote inputs from a variety of brain regions, each with a range of central frequencies above 20Hz. These may be relayed to CA1, possibly via the midline thalamic regions (Jayachandran et al., 2023) or via the entorhinal cortex (Kay & Freeman, 1998; Kay et al., 1996).

There is increasing evidence for the contribution of beta oscillations to a wide range of cognitive processes (Lundqvist et al., 2024; Miles et al., 2023), expanding from prior work focused on sensorimotor control (Barone & Rossiter, 2021; Engel & Fries, 2010) and sensory perception (Kay, 2014). While beta has been implicated in coordinating neural activity across brain networks, including the hippocampus and neocortex (Igarashi et al., 2014; Jayachandran et al., 2023; Spitzer & Haegens, 2017; Symanski et al., 2022), we identified an important period for learning during which the hippocampus and PFC are modulated independently by distinct frequency bands within the beta range. Thus, identifying functional subnetworks and when they coordinate can provide further insight into the temporal dynamics of distributed neural processes supporting cognition.

Methods

Animals

Long-Evans rats (Charles River Laboratories, 6 males, 4-10 months) were used for the study. All procedures were performed under approval by the university’s Institutional Animal Care and Use Committee, according to the guidelines of the Association for Assessment and Accreditation of Laboratory Animal Care.

Neural implants

Animals were chronically implanted with neural recording devices (Voigts et al., 2020) with up to 32 adjustable tetrodes. Up to16 tetrodes targeted the CA1 region of the left hippocampus (AP -3.8, ML -2.75 relative to Bregma), and up to 16 tetrodes targeted both hemispheres of the PFC (AP +3.5, ML +/-1.5 relative to Bregma). The PFC bundle was angled at 15 degrees towards the midline to avoid the superior sagittal sinus. A ground screw was placed in the skull above the right cerebellum. Tetrodes were gradually adjusted into the target regions over two to three weeks after surgery, based on LFP markers and estimated depth of the tetrodes. Once the tetrodes were within the target brain regions, small adjustments of 25μm, when necessary, were made to isolate single cells. Tetrode locations were verified at the end of the experiment. To mark the tip of the tetrode, electrolytic lesions were induced for each tetrode (100μA, 2s). Nissl staining was used to verify the location of the lesion was within the intended area.

Spatial learning apparatus

All tracks and mazes were constructed from acrylic (black DP-9, 8cm wide, 4cm tall walls, 76cm above floor level). Reward delivery in all tasks was automated. Infrared sensors were placed at the ends of the maze arms to detect the rat’s arrival and to trigger the reward. Reward was delivered by a syringe pump (100μl at 20ml/min, NE-500, New Era Pump Systems Inc, New York, USA).

Behavioral training

Animals were food restricted to >85-90% of their baseline body weight. For pretraining, animals were first trained to forage for reward in a black open field box (H: 31cm, W: 61cm, L: 61cm), for 10 minutes per day for 3 days. The reward was 70% evaporated milk (Carnation) with 5% sucrose. The reward was randomly dropped inside the open box to encourage foraging. The next pretraining training phase involved the rats learning to run back and forth on an elevated linear track (60cm) to consume reward from the ends. Animals were trained until a performance criterion of 20 rewards per session (10-minute sessions, three times daily, 4 to 7 days).

For the tasks on the Y and F shaped mazes, the animals needed to learn a specific sequence of goal visits to be rewarded on every trial. There were three goal locations in each maze, which were wells at the end of the arms. The animal needed to alternate visits between wells 2 and 3 via well 1 (1, 2, 1, 3, 1, 2, 1, 3 …), similar to previously described spatial alternation tasks (Frank et al., 2000; Kim and Frank, 2009; Jadhav et al., 2012). For example, if the rat was at well 1, it would be rewarded next at well 2 only if it had previously visited well 3. Similarly, if the rat was at well 1, it would be rewarded next at well 3 only if it had previously visited well 2. If the rat was at wells 2 or 3, it would be rewarded only by going to well 1. Reward delivery was controlled using custom scripts written for the data recording system.

Data recording

Data were collected using the Trodes data acquisition system (SpikeGadgets LLC, California, USA). Neural data were recorded using a 128-channel digitizing headstage. The head stage was tethered via a commutator to the main data acquisition system. The commutator and tether were supported by a passive suspension system. Raw data were recorded at 30KHz. Local field potential (<300Hz) was extracted at 1500Hz. Time-synchronized video (1456 × 1088, 30Hz) was recorded using Allied Vision Manta G-158C cameras.

Behavioral analysis

DeepLabCut (Mathis et al., 2018) was used to track the animal in the video. The spatial coordinate for the neck of the animal was used as the animal’s location and to calculate the animal’s speed. Performance of the animals was derived using a state-space model that estimated the probability of the animal making a correct choice (Smith et al., 2004). We then calculated the mean performance for each day grouped performance by quintiles.

Data and statistical analysis packages

We used Python packages for signal processing: Elephant electrophysiology toolkit (https://elephant.readthedocs.io/), fooof tools (https://fooof-tools.github.io/), GhostiPy (https://github.com/kemerelab/ghostipy); circular statistics: PyCircStat (http://github.com/circstat/pycircstat); and statistics: SciPy (scipy.org), statsmodels (statsmodels.org) and pinguoin (pingouin-stats.org).

Spectral analysis

We calculated the power spectral density (PSD) using the Welch function, with 1s segments with 50% overlap (scipy.signal.welch). We calculated coherence using the coherence function with 1s, with 1s segments and 50% overlap (scipy.signal.coherence). We isolated the periodic component of the PSD by subtracting the aperiodic component using Fitting Oscillations & One-Over-F (fooof-tools) (Donoghue et al., 2020).

Beta burst detection

Local field potential (LFP) analysis was performed using the Elephant electrophysiology toolkit (Denker et al., 2024). We applied a continuous wavelet transform (CWT) using the generalized Morse wavelet (γ = 4, β=7.5) to extract time-resolved spectral power for 0.5–130 Hz in 1 Hz steps. This was done using the Ghostipy package (Chu & Kemere, 2021). The average power in the beta-band (15–30 Hz) isolated from the CWT spectrogram was used to detect bursts using a dual-threshold amplitude method (Feingold et al., 2015). Bursts were defined as periods when the beta power envelope exceeded 1 standard deviation (SD) above the mean power with a minimum duration of 100ms. The onset and offset of each burst were marked by the first upward and downward crossings of 0.5 SD above the mean, respectively. Burst peaks were characterized by the z-scored amplitude at the power maximum. Events separated by less than 100ms were combined. Periods where the maximum LFP signal amplitude across all tetrodes exceeded 10 standard deviations were classified as noise and excluded from analyses.

We identified the time at which the burst had its peak power, the magnitude of the peak power, and the frequency at the time of peak power. This was done by isolating the CWT spectrogram for each burst and extracting the time and frequency corresponding to the maximum power value within the burst. Bursts were classified as independent or coincident based on overlap. Independent bursts are those without overlap in time, whereas coincident bursts overlap in time.

Burst aligned power and coherence analysis

Event-aligned power and coherence within and across PFC and CA1 were computed using the multi-taper method (2 tapers with a time-bandwidth product of 2) (Python implementation of the coherence function from chronux.org) (Mitra & Bokil, 2008). We calculated the coherence in a 1s window centered at the peak of each burst (500ms moving window, 5ms steps). For within-region coherence, the coherence value was the mean for all pairs of tetrodes. To determine the change in coherence around the peak of the burst, a baseline was defined as the mean coherence in the −500 to −375ms interval relative to the center of the burst. We used linear mixed-effects models (statsmodels.mixedlm) to determine the relationship between burst power or frequency variance and performance (by day or quintiles) while accounting for differences between animals.

Sharp wave ripple detection

Sharp-wave ripples (SWRs) were identified based on previous methods (http://github.com/Eden-Kramer-Lab/ripple_detection) (Kay et al., 2016). LFP from tetrodes in the dorsal CA1 region (oriens and pyramidale) was filtered in the ripple frequency band (150–250 Hz). The median of the smoothed envelope (Gaussian kernel σ = 4 ms) was calculated. We used a threshold of 2 standard deviations for detecting SWRs, with a minimum duration of 15ms and separation of 50ms. Only SWR events occurring at speed <4 cm/s and at the goal locations were included. We filtered out detection artifacts due to LFP noise by excluding time intervals in which the LFP signal magnitude exceeded 10 standard deviations across any tetrode.

Spike sorting

We used MountainSort (Chung et al., 2017) with manual curation to isolate single cells. The quality of well-isolated cells was determined by having an interspike interval >2ms, consistency of the waveforms, and the stability of the peak voltage values over the sessions.

Spiking phase locking analysis

For each tetrode, the phase of beta oscillations was extracted by first applying a band-pass filter to the raw LFP signal within the beta band range (15-30Hz, acausal 4th order Butterworth),followed by a Hilbert transform using the Elephant toolkit (Denker et al., 2024). For each cell, we calculated the phase at each spike relative to the beta phase for every available tetrode, which allowed us to determine the cell’s phase-locking preference to beta within the same brain region (local) and the other region (remote). We used two metrics to quantify the strength of spike phase locking to beta. First, we performed the Rayleigh test for circular non-uniformity and calculated the Rayleigh-z value using PyCircStat (Berens, 2009). Second, we fitted a von Mises distribution to the phase preference histogram to estimate the mean phase and the concentration parameter (kappa) (scipy.stats.vonmises). To determine the local phase-locking strength of a given cell, we chose the maximum values of the phase-locking metric across all tetrodes within the same region as the cell. Similarly, we calculated the remote phase-locking strength as the maximum phase locking value across all tetrodes in the other brain region. Only spikes that occurred when the animal was at the goal location were included for the phase-locking analysis.

SWR modulation of spiking

SWR modulation was determined based on previously described methods (Jadhav et al., 2016; Rothschild et al., 2017; Yu et al., 2017; Yu et al., 2018). For each cell, a SWR event-aligned spiking histogram was created for a 1 s window centered on the start of each SWR. A modulation index was calculated based on the difference in spiking rate within a 200 ms window after the start of the SWR relative to the baseline within the 1s window. A circular-shift permutation test was used to determine the significance of the modulation index. A positive modulation index corresponds to an increase in firing around the time of SWRs, whereas a negative modulation index corresponds to a decrease in firing. A permutation test was used to determine the significance.

Correlation between beta phase locking preference and SWR modulation index or goal-entry firing

We calculated the Pearson’s correlation between a cell’s beta phase preference and its SWR modulation index or goal-entry firing. Goal-entry firing measures the firing latency relative to goal entry and is defined as the time corresponding to 50% of the firing density in the 5 s from goal entry. To estimate the significance of the correlation coefficient, we performed a permutation test. Each cell was classified based on its phase-locking significance to local beta (β+ or β-) and its SWR modulation significance (SWR+ or SWR-). This produces four possible combinations in the population (β-SWR-, β+SWR-, β-SWR+, or β+SWR+). We then permuted the β and SWR significance labels for the population and repeated the calculation. This was done 5000 times to generate a permuted correlation coefficient distribution, which allowed us to estimate the significance to p=0.0002. The p value of the observed correlation coefficient is calculated as the proportion of the permuted distribution that is greater or less than the observed value.

eLife Assessment

This paper represents a valuable contribution to our understanding of how LFP oscillations and beta band coordination between the hippocampus and prefrontal cortex of rats may relate to learning. Enthusiasm for the reported results was moderated by the concern that some key analyses need to be done, and highly relevant details about task, data, and statistics were missing. Consequently, the reviewers considered the evidence to be incomplete in this version of the manuscript.

Reviewer #1 (Public review):

Wang, Zhou et al. investigated coordination between the prefrontal cortex (PFC) and the hippocampus (Hp), during reward delivery, by analyzing beta oscillations. Beta oscillations are associated with various cognitive functions, but their role in coordinating brain networks during learning is still not thoroughly understood. The authors focused on the changes in power, peak frequencies, and coherence of beta oscillations in two regions when rats learn a spatial task over days. Inconsistent with the authors' hypothesis, beta oscillations in those two regions during reward delivery were not coupled in spectral or temporal aspects. They were, however, able to show reverse changes in beta oscillations in PFC and Hp as the animal's performance got better. The authors were also able to show a small subset of cell populations in PFC that are modulated by both beta oscillations in PFC and sharp wave ripples in Hp. A similarly modulated cell population was not observed in Hp. These results are valuable in pointing out distinct periods during a spatial task when two regions modulate their activity independently from each other.

The authors included a detailed analysis of the data to support their conclusions. However, some clarifications would help their presentation, as well as help readers to have a clear understanding.

(1) The crucial time point of the analysis is the goal entry. However, it needs a better explanation in the methods or in figures of what a goal entry in their behavioral task means.

(2) Regarding Figure 2, the authors have mentioned in the methods that PFC tetrodes have targeted both hemispheres. It might be trivial, but a supplementary graph or a paragraph about differences or similarities between contralateral and ipsilateral tetrodes to Hp might help readers.

(3) The authors have looked at changes in burst properties over days of training. For the coincidence of beta bursts between PFC and Hp, is there a change in the coincidence of bursts depending on the day or performance of the animal?

(4) Regarding the changes in performance through days as well as variance of the beta burst frequency variance (Figures 3C and 4C); was there a change in the number of the beta bursts as animals learn the task, which might affect variance indirectly?

(5) In the behavioral task, within a session, animals needed to alternate between two wells, but the central arm (1) was in the same location. Did the authors alternate the location of well number 1 between days to different arms? It is possible that having well number 1 in the same location through days might have an effect on beta bursts, as they would get more rewards in well number 1?

(6) The animals did not increase their performance in the F maze as much as they increased it in the Y maze. It would be more helpful to see a comparison between mazes in Figure 5 in terms of beta burst timing. It seems like in Y maze, unrewarded trials have earlier beta bursts in Y maze compared to F maze. Also, is there a difference in beta burst frequencies of rewarded and unrewarded trials?

(7) For individual cell analysis, the authors recorded from Hp and the behavioral task involved spatial learning. It would be helpful to readers if authors mention about place field properties of the cells they have recorded from. It is known that reward cells firing near reward locations have a higher rate to participate in a sharp wave ripple. Factoring in the place field properties of the cells into the analysis might give a clearer picture of the lack of modulation of HP cells by beta and sharp wave ripples.

Reviewer #2 (Public review):

(1) When presenting the power spectra for the representative example (Figure 1), it would be appropriate to display a broader frequency band-including delta, theta, and gamma (up to ~100 Hz), rather than only the beta band. What was the rat's locomotor state (e.g., running speed) after entering the reward location, during which the LFPs were recorded? If the rats stopped at the goal but still consumed the reward (i.e., exhibited very low running speed), theta rhythms might still occasionally occur, and sharp-wave ripples (SWRs) could be observed during rest. Do beta bursts also occur during navigation prior to goal entry? It would be beneficial to display these rhythmic activities continuously across both the navigation and goal entry phases. Additionally, given that the hippocampal theta rhythm is typically around 7-8 Hz, while a peak at approximately 15-16 Hz is visible in the power spectra in Figure 1C, the authors should clarify whether the 22 Hz beta activity represents a genuine oscillation rather than a harmonic of the theta rhythm.

(2) The authors claim that beta activity is independent between CA1 and PFC, based on the low coherence between these regions. However, it is challenging to discern beta-specific coherence in CA1; instead, coherence appears elevated across a broader frequency band (Figure 2 and Figure 2-1D). An alternative explanation could be that the uncoupled beta between CA1 and PFC results from low local beta coherence within CA1 itself.

(3) In Figure 2-1E-F, visual inspection of the box plots reveals minimal differences between PFC-Ind and PFC-Coin/CA1-Coin conditions, despite reported statistical significance. It may be necessary to verify whether the significance arises from a large sample size.

(4) In Figure 3 and Figure 4, although differences in power and frequency appear to change significantly across days, these changes are not easily discernible by visual inspection. It is worth considering whether these variations are related to increased task familiarity over days, potentially accompanied by higher running speeds.

(5) The stronger spiking modulation by local beta oscillations shown in Figure 6 could also be interpreted in the context of uncoupled beta between CA1 and PFC. In this analysis, only spikes occurring during beta bursts should be included, rather than all spikes within a trial. The authors should verify the dataset used and consider including a representative example illustrating beta modulation of single-unit spiking.

(6) As observed in Figure 7D, CA1 beta bursts continue to occur even after 2.5 seconds following goal entry, when SWRs begin to emerge. Do these oscillations alternate over time, or do they coexist with some form of cross-frequency coupling?

Reviewer #3 (Public review):

Summary:

This paper explored the role of beta rhythms in the context of spatial learning and mPFC-hippocampal dynamics. The authors characterized mPFC and hippocampal beta oscillations, examining how their coordination and their spectral profiles related to learning and prefrontal neuronal firing. Rats performed two tasks, a Y-maze and an F-maze, with the F-maze task being more cognitively demanding. Across learning, prefrontal beta oscillation power increased while beta frequency decreased. In contrast, hippocampal beta power and beta frequency decreased. This was particularly the case for the well-performed and well-learned Y-maze paradigm. The authors identified the timing of beta oscillations, revealing an interesting shift in beta burst timing relative to reward entry as learning progressed. They also discovered an interesting population of prefrontal neurons that were tuned to both prefrontal beta and hippocampal sharp-wave ripple events, revealing a spectrum of SWR-excited and SWR-inhibited neurons that were differentially phase locked to prefrontal beta rhythms.

In sum, the authors set out to examine how beta rhythms and their coordination were related to learning and goal occupancy. The authors identified a set of learning and goal-related correlates at the level of LFP and spike-LFP interactions, but did not report on spike-behavioral correlates.

Strengths:

Pairing dual recordings of medial prefrontal cortex (mPFC) and CA1 with learning of spatial memory tasks is a strength of this paper. The authors also discovered an interesting population of prefrontal neurons modulated by both beta and CA1 sharp-wave ripple (SWR) events, showing a relationship between SWR-excited and SWR-inhibited neurons and beta oscillation phase.

Weaknesses:

The authors report on a task where rats were performing sub-optimally (F-maze), weakening claims. Likewise, it is questionable as to whether mPFC and hippocampus are dually required to perform a no-delay Y-maze task at day 5, where rats are performing near 100%. There would be little reason to suspect strong oscillatory coupling when task performance is poor and/or independent of mPFC-HPC communication (Jones and Wilson, 2005), potentially weakening conclusions about independent beta rhythms. Moreover, there is little detail provided about sample sizes and how data sampling is being performed (e.g., rats, sessions, or trials), raising generalizability concerns.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Wang, Zhou et al. investigated coordination between the prefrontal cortex (PFC) and the hippocampus (Hp), during reward delivery, by analyzing beta oscillations. Beta oscillations are associated with various cognitive functions, but their role in coordinating brain networks during learning is still not thoroughly understood. The authors focused on the changes in power, peak frequencies, and coherence of beta oscillations in two regions when rats learn a spatial task over days. Inconsistent with the authors' hypothesis, beta oscillations in those two regions during reward delivery were not coupled in spectral or temporal aspects. They were, however, able to show reverse changes in beta oscillations in PFC and Hp as the animal's performance got better. The authors were also able to show a small subset of cell populations in PFC that are modulated by both beta oscillations in PFC and sharp wave ripples in Hp. A similarly modulated cell population was not observed in Hp. These results are valuable in pointing out distinct periods during a spatial task when two regions modulate their activity independently from each other.

The authors included a detailed analysis of the data to support their conclusions. However, some clarifications would help their presentation, as well as help readers to have a clear understanding.

(1) The crucial time point of the analysis is the goal entry. However, it needs a better explanation in the methods or in figures of what a goal entry in their behavioral task means.

We appreciate Reviewer 1 pointing out this shortcoming and will clarify the description in the revised manuscript. Each goal is located at the end of the arm, and is equipped with a reward delivery unit. The unit has an infrared sensor. The rat breaks the infrared beam when it enters the goal. Figures 1 and 2 have been updated to clearly indicate the time of goal entry. The main text and methods have been updated with the explanation.

(2) Regarding Figure 2, the authors have mentioned in the methods that PFC tetrodes have targeted both hemispheres. It might be trivial, but a supplementary graph or a paragraph about differences or similarities between contralateral and ipsilateral tetrodes to Hp might help readers.

We appreciate this suggestion, which has led to an interesting finding. The coherence and burst coordination were similar for ipsi- and contralateral PFC and hippocampus. Interestingly, we found PFC beta activity was more coherent within each PFC hemisphere compared with across hemispheres. This was observed for coherence and burst time. This suggests there is hemispheric localization of beta oscillations. These results are shown in Fig. 2-1.

(3) The authors have looked at changes in burst properties over days of training. For the coincidence of beta bursts between PFC and Hp, is there a change in the coincidence of bursts depending on the day or performance of the animal?

This is now reported in Fig. 3-3. After quantifying the proportion of independent and coincident bursts as function of experiment day or performance, we found a decrease in the proportion of coincident bursts within CA1, which was specific to the well-performed Y-maze.

(4) Regarding the changes in performance through days as well as variance of the beta burst frequency variance (Figures 3C and 4C); was there a change in the number of the beta bursts as animals learn the task, which might affect variance indirectly?

The difference in the burst count across days did not explain the results. We performed a permutation test (Fig. 4-4), where we randomly shuffled the day identity to control for the count difference across days. The change in variance remains significant.

(5) In the behavioral task, within a session, animals needed to alternate between two wells, but the central arm (1) was in the same location. Did the authors alternate the location of well number 1 between days to different arms? It is possible that having well number 1 in the same location through days might have an effect on beta bursts, as they would get more rewards in well number 1?

The central arm remained the same across days since we needed the animals to learn the alternation task. In our experience, the animal needs a few days to learn the alternation rule when we switch the central arm location. For this experiment, we were interested in the initial learning process, and we kept the central arm constant. Switching the central arm location is a great suggestion for a follow-up experiment where we can understand the effects of reward contingency change on beta bursts.

(6) The animals did not increase their performance in the F maze as much as they increased it in the Y maze. It would be more helpful to see a comparison between mazes in Figure 5 in terms of beta burst timing. It seems like in Y maze, unrewarded trials have earlier beta bursts in Y maze compared to F maze. Also, is there a difference in beta burst frequencies of rewarded and unrewarded trials?

We performed the analysis and found burst timing was similar between the two mazes (Fig. 4-2). Bursts on rewarded trials occurred later than those on unrewarded trials (Fig. 4). Interestingly, PFC bursts on rewarded trials were lower in frequency compared with unrewarded trials. CA1 bursts during rewarded and unrewarded trials had similar frequencies (Fig. 4-3).

(7) For individual cell analysis, the authors recorded from Hp and the behavioral task involved spatial learning. It would be helpful to readers if authors mention about place field properties of the cells they have recorded from. It is known that reward cells firing near reward locations have a higher rate to participate in a sharp wave ripple. Factoring in the place field properties of the cells into the analysis might give a clearer picture of the lack of modulation of HP cells by beta and sharp wave ripples.

As recommended, we quantified the mean speed, mean distance to goal locations, and spatial information for CA1 cells (Fig. 7 J-L). We found SWR reactivated CA1 cells had higher speed and were spiking further away from the goals compared with non-reactivated CA1 cells. This is consistent with prior work that shows SWR-associated reactivation in dorsal CA1 can correspond to trajectories taken as the animal moves towards goals. In intermediate CA1, the content of reactivations is biased toward place representations closer to goals (Jin et al., 2024). CA1 cells with or without phase locking to beta oscillations had similar spatial firing properties.

Reviewer #1 (Recommendations for the authors):

(1) Please make a figure representing what the goal entry means in Figure 1.

We have updated Fig. 1 to clearly show the definition of goal entry. We also edited the main text and methods to better explain the definition of goal entry.

(2) For Figure 1-1, please either change the contrast of the pictures, or define the lesioned areas, as it is a bit difficult to see the lesioned parts, especially in PFC.

We have increased the contrast for Fig. 1-1 for the histology to better show the lesions.

(3) For Figure 1-2, is it possible to show a beta burst from Hp?

Yes, we provided two examples of beta bursts from the hippocampus alongside burst examples from PFC in Fig. 1-2.

(4) Is it possible to make a supplementary table showing the number of tetrodes recorded from each animal per day, plus the number of isolated single cells?

Yes, we have included tetrode counts in Table 1-2 and cell counts in Tables 5-2 to 5-4.

Reviewer #2 (Public review):

(1) When presenting the power spectra for the representative example (Figure 1), it would be appropriate to display a broader frequency band-including delta, theta, and gamma (up to ~100 Hz), rather than only the beta band.

We agree the extended frequency range provides a better overview of the spectral characteristics during the goal period. We have now included example spectrograms up to 100 Hz to show the spectral content for a wider range of frequencies (Fig. 1-2). Further, we have included additional analyses to compare the spectral characteristics between periods when the animal was moving on the maze or immobile at the goal, for frequencies up to 100 Hz (Fig. 1C-H, Fig. 1-3 and 1-4). We used both Welch’s periodogram (Fig. 1-3) and continuous wavelet transform (Fig. 1-4) to demonstrate our findings on beta oscillations in both regions are robust and consistent.

What was the rat's locomotor state (e.g., running speed) after entering the reward location, during which the LFPs were recorded?

Because goal entry is defined as the time the animals break the infrared beam at the goal (response to Reviewer 1), the rat would have come to a stop. We have added the time-aligned speed profile to the spectra and raw data examples in the manuscript (Fig. 1B, Fig. 1-4, Fig. 2A, and Fig. 6A). In addition, we added the quantification of the animal’s speed at the time of beta bursts (Fig. 3-1, Fig. 4-1) and SWRs (Fig. 6-1D).

If the rats stopped at the goal but still consumed the reward (i.e., exhibited very low running speed), theta rhythms might still occasionally occur, and sharp-wave ripples (SWRs) could be observed during rest.

We typically find low theta power in the hippocampus after the animal reaches the goal location and as it consumes reward. Reviewer 2 is correct about occasional theta power at the goal. To compare differences in LFP characteristics between maze running and goal locations, we added additional analyses in Fig. 1-3 and 1-4. We did find SWRs during goal periods (Fig. 6) and we quantified SWR properties in an additional analysis in Fig. 6-1.

Do beta bursts also occur during navigation prior to goal entry? It would be beneficial to display these rhythmic activities continuously across both the navigation and goal entry phases.

We did not find consistent beta bursts in PFC or CA1 on approach to goal entry. We generated an additional goal entry-aligned spectrogram, and quantification (Fig. 1-4) to show that beta oscillations in both regions increased after goal entry. This was also supported by the Welch’s periodogram method (Fig. 1C-H, Fig. 1-3). Beta oscillations in the hippocampus during locomotion or exploration have been reported (Ahmed & Mehta, 2012; Berke et al., 2008; França et al., 2014; França et al., 2021; Iwasaki et al., 2021; Lansink et al., 2016; Rangel et al., 2015).

Additionally, given that the hippocampal theta rhythm is typically around 7-8 Hz, while a peak at approximately 15-16 Hz is visible in the power spectra in Figure 1C, the authors should clarify whether the 22 Hz beta activity represents a genuine oscillation rather than a harmonic of the theta rhythm.

We performed further spectral analysis comparing times when the animal is moving on the maze with times when the animal is immobile at the goal (Fig. 1-3 and 1-4). The results point to the beta frequency oscillations in both regions are unlikely to be harmonics of theta. The beta frequency bands in the spectrogram are independent of the theta band. We were initially concerned about the possibility that the 22 Hz power in CA1 may be a harmonic rather than a standalone oscillation band. If these are harmonics of theta, we should expect to find coincident theta at the time of bursts in the beta frequency. In Fig. 1B, Fig. 1-5, and Fig. 2A, we show examples of the raw LFP traces from CA1. Here, the detected bursts are not accompanied by visible theta-frequency activity. For PFC, we do not always see persistent theta-frequency oscillations like CA1. In PFC, we found beta bursts were frequent and visually identifiable when examining the LFP. We provided examples of the PFC LFP (Fig. 1B, Fig. 1-5, and Fig. 2A). In these cases, we see clear beta frequency oscillations lasting several cycles and these are not accompanied by any visible oscillations in the theta frequency in the LFP trace.

(2) The authors claim that beta activity is independent between CA1 and PFC, based on the low coherence between these regions. However, it is challenging to discern beta-specific coherence in CA1; instead, coherence appears elevated across a broader frequency band (Figure 2 and Figure 2-1D). An alternative explanation could be that the uncoupled beta between CA1 and PFC results from low local beta coherence within CA1 itself.

This is a legitimate concern, and we used three methods to characterize coherence and coordination between the two regions. First, we calculated coherence for tetrode pairs for times when the animal was at goals (Fig. 2B), which provides a general estimation of coherence across frequencies but lack any temporal resolution. Second, we calculated burst-aligned coherence (Fig. 2-2), which provides temporal resolution relative to the burst, but the multi-taper method is constrained by the time-frequency resolution trade-off. Third, we quantified the timing between the burst peaks (Fig. 2D), which described the timing differences but the peaks for the bursts may not be symmetric. Each method has its own caveats, but we drew our conclusion from the combination of results from these three analyses, which pointed to similar conclusions.

Reviewer 2 is correct in pointing out the uniformly high coherence within CA1 across the frequency range we examined. When we inspected the raw LFP across multiple tetrodes in CA1, they were similar to each other (Fig. 2A). This likely reflects the uniformity in the LFP across recording sites in CA1, which is what we saw with coherence values across the frequency range (Fig. 2B). We found that CA1 coherence between tetrode pairs within CA1 was statistically higher than tetrode pairs in PFC across the frequency range (Fig. 2B and C), thus our results are unlikely to be explained by low beta coherence within CA1 itself. The burst-aligned coherence using a multi-taper method also supports this. The coherence values within CA1 at the time of CA1 bursts were ~0.8-0.9.

(3) In Figure 2-1E-F, visual inspection of the box plots reveals minimal differences between PFC-Ind and PFC-Coin/CA1-Coin conditions, despite reported statistical significance. It may be necessary to verify whether the significance arises from a large sample size.

We will include the sample sizes in Table 2-1. We repeated the analyses based on average values per day for each animal (Fig. 2-2 E-F). The pattern of significance remains consistent.

(4) In Figure 3 and Figure 4, although differences in power and frequency appear to change significantly across days, these changes are not easily discernible by visual inspection. It is worth considering whether these variations are related to increased task familiarity over days, potentially accompanied by higher running speeds.

We agree with Reviewer 2 that familiarity increases across days, and the animal is likely running faster. The analysis for Fig. 3 (previously Fig. 3 and 4) includes only data from periods when the animal was at the goal and was not moving. We added a supplemental figure (Fig. 3-1), which shows that the speed of the animal was below 0.5 cm/s at the time of the analyzed bursts. We used linear mixed-effects models to quantify the relationship between power, frequency and day or behavioral quintile, which accounts for repeated measurements across animals.

(5) The stronger spiking modulation by local beta oscillations shown in Figure 6 could also be interpreted in the context of uncoupled beta between CA1 and PFC. In this analysis, only spikes occurring during beta bursts should be included, rather than all spikes within a trial. The authors should verify the dataset used and consider including a representative example illustrating beta modulation of single-unit spiking.

We agree with Reviewer 2 that the stronger modulation to local beta is another piece of evidence indicating uncoupled beta between the two regions. We appreciate this suggestion and have revised Fig. 5 (previously Fig. 6) to include examples illustrating beta modulation for single units. These are spike-phase raster and histograms. We want to clarify that in the revised manuscript, the spikes were only from periods when the animal was at the goal location (5 s after entry) and did not include the running period between goals. Although beta power fluctuates in bursts, our data show there are ongoing beta oscillations throughout this period (Fig. 1-3 and Fig. 1-4), which prompted us to examine the entire period.

(6) As observed in Figure 7D, CA1 beta bursts continue to occur even after 2.5 seconds following goal entry, when SWRs begin to emerge. Do these oscillations alternate over time, or do they coexist with some form of cross-frequency coupling?

This is a very helpful suggestion and led to some interesting findings. We performed two additional analyses: 1) burst/SWR cross-correlation on a shorter timescale and 2) SWR-aligned spectrogram for the beta frequency range. PFC beta burst timing and power appear to be anti-correlated with SWRs detected in CA1 (Fig. 6G and K). In contrast CA1 beta bursts were more likely to occur with SWRs (Fig. 6H and L). These results suggest there is temporal coordination between ongoing beta oscillations in PFC and SWRs in the hippocampus during waking. Cortical beta oscillations are reduced during hippocampal SWRs, perhaps to support the transient switch in global cortical states that accompanies SWRs.

To examine potential cross-frequency coupling between SWRs and beta oscillations, we computed the mean SWR band power (150-250).

Reviewer #3 (Public review):

Summary:

This paper explored the role of beta rhythms in the context of spatial learning and mPFC-hippocampal dynamics. The authors characterized mPFC and hippocampal beta oscillations, examining how their coordination and their spectral profiles related to learning and prefrontal neuronal firing. Rats performed two tasks, a Y-maze and an F-maze, with the F-maze task being more cognitively demanding. Across learning, prefrontal beta oscillation power increased while beta frequency decreased. In contrast, hippocampal beta power and beta frequency decreased. This was particularly the case for the well-performed and well-learned Y-maze paradigm. The authors identified the timing of beta oscillations, revealing an interesting shift in beta burst timing relative to reward entry as learning progressed. They also discovered an interesting population of prefrontal neurons that were tuned to both prefrontal beta and hippocampal sharp-wave ripple events, revealing a spectrum of SWR-excited and SWR-inhibited neurons that were differentially phase locked to prefrontal beta rhythms.

In sum, the authors set out to examine how beta rhythms and their coordination were related to learning and goal occupancy. The authors identified a set of learning and goal-related correlates at the level of LFP and spike-LFP interactions, but did not report on spike-behavioral correlates.

Strengths:

Pairing dual recordings of medial prefrontal cortex (mPFC) and CA1 with learning of spatial memory tasks is a strength of this paper. The authors also discovered an interesting population of prefrontal neurons modulated by both beta and CA1 sharpwave ripple (SWR) events, showing a relationship between SWR-excited and SWR-inhibited neurons and beta oscillation phase.

Weaknesses:

Moreover, there is little detail provided about sample sizes and how data sampling is being performed (e.g., rats, sessions, or trials), raising generalizability concerns.

We appreciate Reviewer 3’s thoughtful suggestions for making our claims convincing. We have included information about sample sizes in the revised manuscript.

The authors report on a task where rats were performing sub-optimally (F-maze), weakening claims.

Our experiment was designed to create a scenario in which one task was learned (Y-maze) and another was not (F-maze). This contrast allows us to determine differences in neural correlates of learning versus familiarity. The design produced a learned and not learned task with similar levels of familiarity over 5 days, within the same animal.

Likewise, it is questionable as to whether mPFC and hippocampus are dually required to perform a no-delay Y-maze task at day 5, where rats are performing near 100%.

We agree with Reviewer 3 that the mPFC and hippocampus may not be required when the animal reaches stable performance on day 5 (Deceuninck & Kloosterman, 2024). The data we collected spans the full range of early learning (day 1) to proficiency (day 5). We wanted to understand the dynamics of beta across these learning stages, which have not been reported previously.

Recent studies suggest mPFC and hippocampus are likely to be needed, in some capacity, for learning continuous spatial alternation tasks on a range of maze geometries. Lesions, inactivation or waking activity perturbation of hippocampus or hippocampus and mPFC on the W maze alternation task slowed learning (Jadhav et al., 2012; Kim & Frank, 2009; Maharjan et al., 2018). More recently, optogenetic silencing of mPFC after sharp wave ripples on the Y-maze alternation affected performance when the center arm was switched (den Bakker et al., 2023). The Y and F-mazes in our study both share the continuous alternation rule, where the animal needed to avoid visiting a previously visited location on the outbound choice relative to the center, and always return to the center location.

Further, the performance characteristics on the outbound and inbound components of our Y task are similar to the W task. We have analyzed the “inbound” and “outbound” performance of the animals on the Y-maze alternation task, and they are similar to the W maze alternation task. The “inbound” or reference location component is learned quickly whereas the “outbound”, alternation component is learned slowly.

There would be little reason to suspect strong oscillatory coupling when task performance is poor and/or independent of mPFC-HPC communication (Jones and Wilson, 2005) potentially weakening conclusions about independent beta rhythms.

Although many studies have examined the oscillatory coupling properties at the theta frequency between mPFC-HPC (Hyman et al., 2005; Jones & Wilson, 2005; Siapas et al., 2005), our understanding of beta frequency coordination between the two regions is less established, especially at goal locations. Our work suggests beta frequency coordination at goal locations does not share properties with those of theta frequency coupling between mPFC and HPC, which occurs primarily during movement on the maze. Our first novel finding is that beta oscillations occur at goal locations in mPFC and HPC; our second is that the beta frequency dynamics in these regions are surprisingly distinct. We are not aware of prior work describing these properties at goal locations in spatial navigation tasks, especially their temporal coordination.

Reviewer #3 (Recommendations for the authors):

(1) The conclusions from this article would be made much stronger if the authors (1) record from rats performing a task known to be dependent on mPFC-HPC communication (e.g. a spatial working memory task) or (2) record from the F-maze in well-trained rats (75-80% performance is common), or (3) show that Y-maze task performance is dependent on mPFC-hippocampal communication (see Maharjan et al., 2018, which used a W-track). It is possible that learning the Y-maze depends on mPFC-hippocampal communication, but that with asymptotic performance, this changes. This would put beta oscillation coupling findings into a nuanced perspective.

We appreciate the recommendation. The objective of this current manuscript is to present previously unknown properties of beta oscillations in hippocampal-prefrontal cortical networks. We agree that further investigation is required to fully dissect the functional contribution of beta dynamics in these networks. We are in the process of doing that.

The rule on the Y-maze in our experiment is identical to the continuous alternation rule on the W maze in Maharjan et al., 2018 and Kim and Frank, 2009 (Author response image 1), which showed the PFC and hippocampus are required for normal learning, respectively. The Y and W mazes share the same topology; there is one junction connecting three arms. The W maze has two 90-degree-angle turns which are not choice points. Thus, both maze tasks have one choice point and involve learning an alternation rule. Further the learning properties share similarities. For the Y-maze, the inbound portion (return to center) (Author response image 1) was quicker to learn than the outbound portion (alternation) (Author response image 1). The same pattern is observed in the W maze learning task (Maharjan et al., 2018, Fig. 3A and D, Kim and Frank, Fig. 4C). We agree an experiment is needed to show that Y-maze task learning is dependent on the function of hippocampal-prefrontal cortical networks. The shared topology, rule definition, and behavior profile suggest that the Y and W maze tasks engage similar learning processes.

W and Y maze alternation tasks share the same rule.

Schematics illustrate a comparison of W and Y maze alternation task rules. The W maze has been inverted for visual comparison with the Y maze. Performance grouped by in- or outbound trials. Inbound trials originate from the side goals (2 or 3). Outbound trials originate from the center goal (1). Performance on the inbound trials was higher than that on the outbound trials, which is comparable to previously published alternation tasks on W-shaped mazes.

(2) Typically, when analyzing LFP profiles, experimenters include running velocity/speed. It should be ruled out whether spectral changes are confounded in any way by speed or time spent in the goal zones.

We appreciate this suggestion for better conveying our definition of goal period. This was raised by other reviewers. We have now included the speed profiles for the goal-entry-aligned spectrograms (Fig. 1B, 1-4, 2A, and 6A), as well as the speed quantification at the times of bursts (Fig. 3-1, Fig. 4-1) and SWRs (Fig. 6-1 D). We strictly define the goal period after entry into the sensor on the reward delivery device at the end of the arms. This is to ensure the animal is immobile for the period to avoid confounds related to speed. We also ensured the time periods are comparable between the trials we analyzed.

(3) The authors should describe in each analysis how data are sampled, and for extracellular electrophysiology experiments, cell counts from each rat reported. A table would be ideal. For example, it is unknown if entrainment analysis is performed on data collected from one rat or from all rats.

We have now added the missing information (Table 1-2, 5-1 to 5-3). We performed analyses using data from all animals.

(4) There was no profiling of mPFC neurons in terms of their behavioral correlates. The authors should strongly consider examining how individual neurons encode task variables (e.g., trial correctness, reward location...) and can do so using a generalized linear model. Adding an analysis of behavioral correlates could nicely tie into the beta-SWR analyses. For example, are SWR-beta rhythm-modulated neurons also behaviorally modulated?

This is a very helpful suggestion. We have added analyses on the behavioral correlates of PFC and CA1 neurons. We quantified three metrics: 1) whether PFC and CA1 spiking activity can distinguish goal location based on firing rate or phase preference, 2) firing distance to goal and 3) spatial information (Fig. 5 and 7). We also performed the same analysis based on SWR and beta modulation status. We found PFC cells that were both SWR- and beta-modulated showed the strongest task firing relationship (Fig. 7).

(5) Figure 2:

Are these data analyzed from well-trained rats? What is your N (rats/sessions/trials/epochs)?

These results in Fig. 2 are from all days. We added the breakdown in Table 2-1.

(6) Figure 3:

(a) The authors show that on the Y-maze, performance, beta oscillations power, and beta oscillation frequency change over days. However, for the F-maze, performance improves but appears to taper off at 60% and PFC beta frequencies do not change with learning. Do you have rats performing this task well above chance (e.g., 75-80%?), and if so, do beta oscillation frequencies in the mPFC gradually change?

We did not observe rats performing above chance on the F-maze over 5 days. They do not appear to learn the alternation rule on this task. This is why we used the F-maze as the “non-learner” control. The F-maze task design for this study appears to be difficult to learn and would likely require a much longer training period for the performance to exceed chance. We agree an important follow-up question is whether the effects reported here are generally observed across learning in different tasks. This is a future direction we are actively pursuing.

One existing data point may partially address the relevance of beta power change with learning. We analyzed the power as a function of performance quintile on the Y-maze (learned) or F-maze (not learned). The power changed with performance quintile on the Y-maze (learned) but not the F-maze (not learned) (Fig. 3D), pointing to the change in power being associated with learning status.

(b) Does the proportion of beta bursts change with learning? What about the proportion of coherent events?

We performed this analysis and the results are shown in Fig. 3-3. We found a decrease in the proportion of coincident bursts within CA1, which was specific to the well-performed Y-maze.

(c) Is the reduction in beta oscillation frequency in the mPFC related to running behavior? This should be ruled out. Same for the hippocampus.

The reduction is not due to running because we only included bursts when the animal was at the goal and immobile. We added a figure to show speed at the time of bursts (Fig. 3-1).

(d) Why don't you also show coherence as a function of learning?

This is a great suggestion. Coherence as a function of day or performance is now shown in Fig. 3-2. Overall, the trends were weak suggesting there was no strong change in coherence over days or as a function of learning. Although some of the linear mixed effects models were statistically significant, the marginal R2 values (R2m) were very low, indicating the effects of performance or day on coherence were small. Beta frequency coherence within each brain region either remained the same or slightly decreased across days (Fig. 3-2 D and F) or with performance (Fig. 3-2 J). For coherence across regions, we only found a weak but significant increase for the well-performed Y-maze task across days (Fig. 3-2 B).

(e) The statistics in the caption are great, but maybe consider using a table and a supplemental figure

We have now included the most relevant model output in the figure. These are the marginal R2 (R2m) and conditional R2 (R2c). These describe the contribution of the fixed or fixed and random effects on the model, respectively.

(7) Figure 4

(a) This figure is better suited as supplemental to Figure 3.

We have combined Fig. 4 with Fig. 3.

(b) Statistics might be better suited in a table and a supplemental figure.

We have now included the most relevant model output in the figure. These are the marginal R2 (R2m) and conditional R2 (R2c). These describe the contribution of the fixed or fixed and random effects on the model, respectively.

(8) Figure 5

(a) The shift of beta rhythms being linked to learning is interesting, albeit confusing, given that error trials were accompanied by even more 'precise' beta rhythm timing. Is it possible that beta rhythms are accompanied by an error signal? Or maybe instead related to running behavior?

We verified this observation was not due to speed since we only included periods when the animal was at the goal and immobile. We added a new supplemental figure (Fig. 4-1) to report this analysis.

(b) Changes in beta rhythm timing in the F-maze could be entirely explained by the data in the Y-maze, given that F-maze performance was, in general, very low. To test if beta rhythm shifting is a consequence of learning, the authors should show performance on the F-maze without the Y-maze, and vice versa.

As suggested, we also analyzed frequency variance data from each maze separately and found learning-related changes were more prominent on the Y-maze (learned) than on the F-maze (not learned) (Fig. 4-4 C). The F-maze data serves as a valuable control within the same animal for the learning-related effects we are reporting.

We agree that the Y-maze data is consistent with performance-related changes to beta timing. The addition of the F-maze provides a within-animal comparison for a task in which the animal performed less well. We agree an additional control group would provide further evidence to support learning-related changes in beta timing. However, the F-maze will likely take much longer to learn, therefore the additional days of exposure on the task will be a different confound. Thus, we wanted to ensure that we compared a “learned” with a “not learned” task within the same animal, with a comparable exposure period.

(9) Figure 6

(a) How many cells are analyzed here? Are they all pyramidal neurons? The authors should add the number of cells analyzed from each rat. A table would suffice.

We added tables (Table 5-2 to 5-4) to show this information. For our analysis, we included all units.

(b) The authors should include statistics about recorded units. Peak-to-trough timing and interspike interval are commonly used. Even demonstration of action potentials from individual cells is valuable for proof of concept.

We have added more information on the cell type classification (Table 5-2) and how we classified the cells (Methods) and also added example waveforms in Fig. 5 (previously Fig. 6).

(c) Why did the authors stop comparing the F-maze and Y-maze for entrainment analysis?

The data for each maze was comparable. Per the reviewer’s suggestion, we added the entrainment analysis for each maze separately (Fig. 5-1).

(10) Figure 7

(a) The authors should minimally include instantaneous velocity to show that during putative ripple events, the animal is in fact, quiescent.

We added the speed at the time of ripples in Fig. 6-1 D.

(b) Figure 7C: The authors should show how time spent in the reward zones varies over days. Is SWR power simply changing due to behavioral occupancy differences (e.g., a statistical sampling problem)? An analysis of the SWR rate over days would be valuable.

We agree and have added additional SWR properties across days in Fig. 6-1. This includes SWR power (Fig. 6-1 A), rate (Fig. 6-1 B), duration (Fig. 6-1 C) and speed (Fig. 6-1 D). To ensure we are comparing the equivalent time spent at goals, the SWR properties were calculated from the 10 s after goal entry (Fig. 6-1) and we only included goal visits that lasted at least 10 s. This controls for the potential confounds from occupancy differences across days.

(c) Why did the authors stop comparing the F-maze and Y-maze for SWR analysis?

The SWR properties were similar in both tasks. We have now added separate analyses for each maze in Fig. 6-1 and 6-2.

(11) Figure 8

(a) The authors discovered that mPFC neurons more strongly entrained to SWRs compared to their own beta rhythms, potentially indicating coordination of mPFC and hippocampus (lines 275:276). What about mPFC spiking to CA1 beta?

We quantified mPFC spiking entrainment to CA1 beta in Fig. 5 O and P. We found mPFC cells were more strongly entrained to the local beta within mPFC compared with CA1 beta.

(b) Lines 277:278: How did the authors arise at 8.6% being the expected proportion of neurons entrained by both SWRs and beta rhythms?

We have added a better explanation of how we arrived at the expected proportion. The calculation is based on the joint probability between the proportion of neurons modulated by SWRs (30%) and the proportion of neurons with phase locking to beta rhythms (22%). The expected proportion (6.6%) is calculated by multiplying the two values (30% ´ 22%) under the assumption that the two classes are independent. Deviations from the expected proportions are determined using a Fisher’s exact test.

We note that in the revised manuscript we reanalyzed the data with more stringent criteria. We now include SWRs with durations greater than 50 ms, rather than 15 ms, and we quantify beta spike phase-locking using only the spikes within the first 5 s after goal entry, for goal visits lasting at least 5 s.

With these more conservative criteria, the proportion of SWR- and beta-modulated cells (8%) is no longer significantly different from the expected proportion (6.6%, Fisher’s exact test p=0.09). Although this differs from our original significance test, the trend, and importantly, the distinct task correlates for this population (Fig. 7 C-D) still hold. We have updated the revised manuscript with these findings.

Our original result was: “The subset of PFC cells that are modulated by both SWR and beta (11%) is greater than the expected proportion (8.6%) under the assumption that SWR and beta can modulate the population independently (Fisher exact test, p=0.021), although the size of the difference is small.”

(c) The discovery that neurons modulated by both beta and SWRs are unique from those simply modulated by beta is really interesting. The authors discovered an interesting relationship among dually entrained mPFC neurons whereby SWR-excited mPFC neurons were entrained to the peak of beta, whereas SWR-inhibited mPFC neurons were entrained to the trough of beta. Then, in lines285:286, the authors write:

(12) "This relationship was not observed for PFC cells that are modulated by beta but not modulated by SWRs (Fig. 8C, right, Fig. 8-1A)."

Why would the authors expect this relationship to exist when the mPFC neurons were not SWR modulated? Were the authors referring to something else?

We should have phrased this more clearly. We edited the text to better convey the expected SWR and beta modulation patterns for the control populations. We wanted to ensure the relationship between SWR modulation direction and beta phase preference was not observed for the cells that were not SWR-modulated.

Furthermore, what about mPFC neurons that are SWR modulated but not beta modulated? For completeness, the authors should examine these.

We agree this is an important comparison and have now included this population in the new Fig. 7. As expected, we do not find any relationship between SWR modulation direction and spike preference to beta for the SWR-modulated and not beta-modulated population.

(13) Lines 294:296: The authors should elaborate on how they obtained an expected proportion of CA1 beta modulated neurons.

We have now added a description for calculating the expected number of beta-modulated CA1 neurons. These are the neurons that have a Rayleigh test p-value less than 0.05.

(14) What are these neurons doing to predict task information? Do they at all? Do they differ from beta-only neurons or non-phase-locked neurons?

This is a great suggestion, and we have added analyses to show the task correlates for the cells. We found brain region-specific differences for task correlates that depended on how the cell was modulated by SWRs or local beta oscillations. This is shown in the updated Fig. 7 and the accompanying supplemental Figs. 7-1 to 7-2.

For PFC cells, SWR and beta modulation status defined a subpopulation with a strong task structure correlation. There was a positive correlation between the direction of SWR modulation and the spiking distance relative to goals. SWR-excited cells were more active further away from goals, whereas the SWR-inhibited cells were active closer to goals. This correlation was not found for SWR-modulated PFC cells that were not beta-modulated. For PFC cells, the direction of SWR modulation is known to be correlated with movement speed, consistent with the hypothesis that movement-active cells become reactivated during SWRs, and immobility-active cells become suppressed (Jadhav et al., 2016; Yu et al., 2017). We found the same relationship, the direction of SWR modulation was positively correlated with mean spiking speed. Our results show beta modulation marks a subpopulation of PFC cells with stronger task-structure correlates.

For CA1 cells, we found the expected relationship between SWR modulation and task-structure correlates. SWR-excited CA1 cells spiked further away from the goal and when the animal was moving. This is consistent with the reactivation of trajectory-related spatial firing patterns during movement on the maze. This pattern was observed irrespective of beta modulation status.

Ahmed, O. J., & Mehta, M. R. (2012). Running speed alters the frequency of hippocampal gamma oscillations. J Neurosci, 32(21), 7373-7383. https://doi.org/10.1523/JNEUROSCI.5110-11.2012

Berke, J. D., Hetrick, V., Breck, J., & Greene, R. W. (2008). Transient 23-30 Hz oscillations in mouse hippocampus during exploration of novel environments. Hippocampus, 18(5), 519-529. https://doi.org/10.1002/hipo.20435

Deceuninck, L., & Kloosterman, F. (2024). Disruption of awake sharp-wave ripples does not affect memorization of locations in repeated-acquisition spatial memory tasks. Elife, 13. https://doi.org/10.7554/eLife.84004

den Bakker, H., Van Dijck, M., Sun, J. J., & Kloosterman, F. (2023). Sharp-wave-ripple associated activity in the medial prefrontal cortex supports spatial rule switching. Cell Rep, 42(8), 112959. https://doi.org/10.1016/j.celrep.2023.112959

França, A. S., do Nascimento, G. C., Lopes-dos-Santos, V., Muratori, L., Ribeiro, S., Lobão-Soares, B., & Tort, A. B. (2014). Beta2 oscillations (23-30 Hz) in the mouse hippocampus during novel object recognition. Eur J Neurosci, 40(11), 3693-3703. https://doi.org/10.1111/ejn.12739

França, A. S. C., Borgesius, N. Z., Souza, B. C., & Cohen, M. X. (2021). Beta2 Oscillations in Hippocampal-Cortical Circuits During Novelty Detection. Front Syst Neurosci, 15, 617388. https://doi.org/10.3389/fnsys.2021.617388

Hyman, J. M., Zilli, E. A., Paley, A. M., & Hasselmo, M. E. (2005). Medial prefrontal cortex cells show dynamic modulation with the hippocampal theta rhythm dependent on behavior. Hippocampus, 15(6), 739-749. https://doi.org/10.1002/hipo.20106

Iwasaki, S., Sasaki, T., & Ikegaya, Y. (2021). Hippocampal beta oscillations predict mouse object-location associative memory performance. Hippocampus, 31(5), 503-511. https://doi.org/10.1002/hipo.23311

Jadhav, S. P., Kemere, C., German, P. W., & Frank, L. M. (2012). Awake hippocampal sharp-wave ripples support spatial memory. Science (New York, N.Y.), 336(6087), 1454-1458. https://doi.org/10.1126/science.1217230

Jadhav, S. P., Rothschild, G., Roumis, D. K., & Frank, L. M. (2016). Coordinated Excitation and Inhibition of Prefrontal Ensembles during Awake Hippocampal Sharp-Wave Ripple Events. Neuron, 90(1), 113-127. https://doi.org/10.1016/j.neuron.2016.02.010

Jin, S. W., Ha, H. S., & Lee, I. (2024). Selective reactivation of value- and place-dependent information during sharp-wave ripples in the intermediate and dorsal hippocampus. Sci Adv, 10(32), eadn0416. https://doi.org/10.1126/sciadv.adn0416

Jones, M. W., & Wilson, M. A. (2005). Theta Rhythms Coordinate Hippocampal–Prefrontal Interactions in a Spatial Memory Task. PLoS Biology, 3(12). https://doi.org/10.1371/journal.pbio.0030402

Kim, S. M., & Frank, L. M. (2009). Hippocampal Lesions Impair Rapid Learning of a Continuous Spatial Alternation Task. PLoS ONE, 4(5). https://doi.org/10.1371/journal.pone.0005494

Lansink, C. S., Meijer, G. T., Lankelma, J. V., Vinck, M. A., Jackson, J. C., & Pennartz, C. M. (2016). Reward Expectancy Strengthens CA1 Theta and Beta Band Synchronization and Hippocampal-Ventral Striatal Coupling. J Neurosci, 36(41), 10598-10610. https://doi.org/10.1523/JNEUROSCI.0682-16.2016

Maharjan, D. M., Dai, Y. Y., Glantz, E. H., & Jadhav, S. P. (2018). Disruption of dorsal hippocampal-prefrontal interactions using chemogenetic inactivation impairs spatial learning. Neurobiol Learn Mem, 155, 351-360. https://doi.org/10.1016/j.nlm.2018.08.023

Rangel, L. M., Chiba, A. A., & Quinn, L. K. (2015). Theta and beta oscillatory dynamics in the dentate gyrus reveal a shift in network processing state during cue encounters. Front Syst Neurosci, 9, 96. https://doi.org/10.3389/fnsys.2015.00096

Siapas, A. G., Lubenov, E. V., & Wilson, M. A. (2005). Prefrontal Phase Locking to Hippocampal Theta Oscillations. Neuron, 46(1), 141-151. https://doi.org/10.1016/j.neuron.2005.02.028

Yu, J. Y., Kay, K., Liu, D. F., Grossrubatscher, I., Loback, A., Sosa, M.,…Frank, L. M. (2017). Distinct hippocampal-cortical memory representations for experiences associated with movement versus immobility. Elife, 6, e27621. https://doi.org/10.7554/eLife.27621

Version 2

Abstract

Neural activity at goal locations contributes to learning by providing feedback on the success of preceding actions. This period engages neocortical and hippocampal networks, which serve distinct functions in reward processing and in forming associations between experience and reward. A neocortical network signature for reward feedback processing is beta oscillations (15-30 Hz). Beta oscillations are thought to coordinate distributed neural processes across brain regions. However, it is unknown whether beta oscillations coordinate hippocampal-neocortical networks during the goal period, or how their dynamics relate to learning. Here, we show that beta oscillations occur in both hippocampal CA1 and the prefrontal cortex (PFC) when rats reach goal locations in spatial navigation tasks. Despite the presence of beta oscillations in both regions after goal entry, beta activity in each region differed in spectral and temporal properties. These differences suggest that the hippocampus and PFC are weakly coupled at the beta frequency. We found that across learning, the strengths of PFC and CA1 beta oscillations were inversely related: PFC beta power increased and CA1 beta power decreased. Beta burst properties in PFC also had an inverse relationship to those of hippocampal sharp wave-ripples (SWRs), a prominent hippocampal process required for learning. We found a subset of PFC neurons modulated by both beta and hippocampal SWRs, which had distinct task-related firing patterns. Our results suggest that during outcome processing at goal locations, the neocortex and hippocampus are locally modulated at the beta frequency and then become coordinated for memory-related processes during SWRs.

Introduction

Learning leads to changes in distributed neural networks across the brain. During learning, diverse neural processes across brain networks need to be coordinated, which is hypothesized to rely on neural oscillations that rhythmically couple activity in connected brain regions. This coupling between groups of brain regions is thought to be dynamic, creating functionally interacting subnetworks to support specific neural processes when required. A neural oscillation that is hypothesized to support these functions is beta (15-30 Hz) (Bibbig et al., 2002; Gourévitch et al., 2010; Igarashi et al., 2014; Jayachandran et al., 2023; Kopell et al., 2011; Lundqvist et al., 2024; Miles et al., 2023; Spitzer & Haegens, 2017; Symanski et al., 2022). Beta oscillations occur in many brain regions, including sensory (Gervais et al., 2007; Kay et al., 1996; Martin et al., 2004), neocortical (Baker et al., 1997; Buschman et al., 2012; Cohen et al., 2007; Cohen et al., 2015; Dubey et al., 2023; HajiHosseini & Holroyd, 2015; HajiHosseini et al., 2020; Iturra-Mena et al., 2023; Kawasaki & Yamaguchi, 2013; Koloski, Hulyalkar, et al., 2024; MacKay & Mendonça, 1995; Pesaran et al., 2008; Voloh et al., 2020; Walsh et al., 2025; Xiao et al., 2024), subcortical (DeCoteau et al., 2007; Feingold et al., 2015; Howe et al., 2011; Lansink et al., 2009; Leventhal et al., 2012; Quinn et al., 2010; Sturman & Moghaddam, 2012) and limbic regions (Amaya et al., 2024; Berke et al., 2008; Fourcaud-Trocmé et al., 2019; França et al., 2014; França et al., 2021; Igarashi et al., 2014; Jackson et al., 2024; Jayachandran et al., 2023; Kay & Freeman, 1998; Rangel et al., 2015; Rangel et al., 2016; Samson et al., 2017; Symanski et al., 2022). Beta oscillations are associated with a wide range of cognitive functions, including sensory perception (Kay, 2014), reward processing (Cohen et al., 2007), top-down control (Engel & Fries, 2010) and novelty detection (Marco-Pallarés et al., 2015). Despite their widespread occurrence and association with diverse cognitive functions, a less-understood aspect of beta oscillations is their contribution to modulating or coordinating brain networks during learning, when rules are not fully understood and performance is improving. Most studies investigating beta dynamics across the brain have primarily focused on performance after initial learning. However, three lines of work in sensory-limbic, striatal, and amygdala networks point to beta playing a role in supporting learning. First, sensory-association learning studies suggest that beta activity coordinates sensory-limbic networks to enable the formation of associations between the sensory cue and its behavioral relevance (Fourcaud-Trocmé et al., 2019; Martin et al., 2007; Martin et al., 2004). Second, beta oscillation dynamics in the ventral striatum change as habits are formed (Howe et al., 2011). Lastly, beta oscillations in the basolateral amygdala can drive reward preference (Amaya et al., 2024). However, the patterns of learning-related changes in beta oscillations across brain networks remain unclear, since the dynamics of beta during learning outside of olfactory and striatal networks are not known. Here, we asked how beta dynamics contribute to spatial learning in hippocampal-neocortical networks, which are brain regions known to play central roles in supporting memory processing and guiding behavior.

To understand how beta oscillations in hippocampal-neocortical networks change with learning, we focus on an important period for driving learning: when the reward outcome is revealed. We refer to this period as the “goal” phase of a task. Beta oscillations have been observed during equivalent task periods in both human and non-human animal studies. Neocortical beta oscillations during this period have been associated with processing outcomes, where feedback from the presence or absence of the reward can inform future actions and drive learning (Cohen et al., 2011). Goal-related beta oscillations have been observed in frontal cortical regions (Cohen et al., 2007; Kawasaki & Yamaguchi, 2013), including the anterior cingulate cortex (ACC) (Iturra-Mena et al., 2023; Xiao et al., 2024), the prefrontal cortex (PFC) (HajiHosseini & Holroyd, 2015; HajiHosseini et al., 2020; Voloh et al., 2020) and the orbitofrontal cortex (OFC) (Koloski, Hulyalkar, et al., 2024; Koloski, O’Hearn, et al., 2024), in primate and rodent species. Beta oscillations in these frontal cortical regions have been hypothesized to provide top-down cortical control of connected networks (Babapoor-Farrokhran et al., 2017; Benchenane et al., 2011; Bressler & Richter, 2015; Brovelli et al., 2004; Buschman et al., 2012; Dubey et al., 2023; Engel & Fries, 2010; Pesaran et al., 2008).

Despite frontal cortical beta activity being implicated in outcome feedback processing, whether frontal cortical and hippocampal networks are coordinated in the beta frequency range during the goal period remains unknown. Beta frequency coordination of hippocampal-neocortical networks has been reported in olfactory sensory decision-making tasks during the cue presentation period (Igarashi et al., 2014; Jayachandran et al., 2023; Martin et al., 2007; Symanski et al., 2022), and is hypothesized to support the use of memory-based sensory associations for decision making (Spitzer & Haegens, 2017). However, goal-related beta frequency oscillations have not been reported in the hippocampus. Beta oscillations in the hippocampus have so far been observed during exploration (Ahmed & Mehta, 2012), especially in novel environments (Berke et al., 2004), interaction with novel objects (França et al., 2014; França et al., 2021; Rangel et al., 2015), approaches to goals (Iwasaki et al., 2021; Lansink et al., 2009) as well as sensory stimulus sampling (Fourcaud-Trocmé et al., 2019; Gourévitch et al., 2010; Jayachandran et al., 2023; Kay & Freeman, 1998; Kay et al., 1996; Martin et al., 2007; Rangel et al., 2016; Symanski et al., 2022). Thus, it is unknown if and how hippocampal-neocortical networks are coordinated at the beta frequency during the goal phase of a task, and how this coordination relates to learning.

To understand the relationship between goal-related beta frequency activity in hippocampal-neocortical networks and learning, we recorded local field potentials (LFPs) and single units from the hippocampus (dorsal CA1) and the prefrontal cortex (PFC) in rats learning spatial navigation tasks over five days. We observed beta oscillations in both regions at the goal locations of the tasks. However, we were surprised to find that beta oscillations in the PFC and CA1 had distinct spectral and temporal properties. The mismatch in frequency range and timing is consistent with each region engaging in distinct patterns of beta activity, likely with other brain regions, rather than being coupled with each other by beta. Over days of training, beta oscillations in the PFC and CA1 had distinct correlations with performance, where PFC beta increased in strength and CA1 beta decreased in strength. We also found an inverse relationship between the learning-related properties of PFC beta and a prominent hippocampal marker of learning, the sharp wave-ripple (SWR). Lastly, we found distinct task correlates in a subpopulation of PFC cells that showed modulation by both beta oscillations and hippocampal SWRs.

In contrast to previous results demonstrating coordinated hippocampal-neocortical beta modulation during sensory decision-making (Gourévitch et al., 2010; Jayachandran et al., 2023; Kay & Freeman, 1998; Kay et al., 1996; Martin et al., 2007; Rangel et al., 2016; Symanski et al., 2022), we show that beta oscillations in these networks are likely modulating distinct network processes during the goal phase of a task. Further, beta dynamics in the PFC and CA1 undergo opposite changes across learning. After goal entry, the neocortex and hippocampus are weakly coupled at the beta frequency but later coordinate during hippocampal SWRs. Our results support the hypothesis that at the goal, the neocortex is initially engaged in processing outcome-related information before coordinating with hippocampal networks for memory-related processes.

Results

Rats (n=6) were trained on two spatial alternation tasks, interleaved across sessions, for five days (Table 1-1). In each task, reward was delivered when the three goal locations were visited in an alternating sequence (Fig. 1A). We defined the goal location as the reward delivery device at the end of each maze arm (Fig. 1B). Goal entry was defined as the time when the animal arrived at the goal and triggered the infrared sensor on the device. We first asked whether beta oscillations could be observed in both CA1 and PFC (Fig. 1-1 and Table 1-2). We observed power in the beta-frequency range (15-30 Hz) in both PFC and CA1 after the animal reached the goal locations (Fig. 1B, Fig. 1-2). This was confirmed by comparing the power spectral density (PSD) between periods when the animal was moving on the maze arms (On Maze) or immobile at the goal location (At Goal) (Fig. 1C-H and Fig. 1-3). The PSD for PFC showed a peak at approximately 20 Hz (Fig. 1C), which was significantly stronger at the goal compared with on the maze (Fig. 1E and G). In CA1, beta-frequency power was elevated during periods at the goal compared with on the maze (Fig. 1D, F and H). We repeated these analyses using continuous wavelet transform and found the same pattern: beta power was elevated after but not before goal entry (Fig. 1-4).

Distinct PFC and CA1 beta frequency activity after the reward location entry during spatial navigation tasks.

A. Spatial learning tasks on Y and F shaped mazes with spatial alternation rules. Animals were trained on the mazes across interleaved sessions. Reward locations (1, 2 and 3) are indicated. The order of visits required to ensure reward on every trial is shown above the maze schematics. The daily training schedule shows the duration, order and separation between sessions. B. Example PFC (orange) and CA1 (blue) LFP traces aligned to goal entry (black vertical line). Goal entry is the time when the animal reaches the reward delivery device at the end of each maze arm and breaks the infrared beam. This time is registered by an infrared sensor on the device. The beta frequency-filtered LFP is shown below the raw LFP. The heatmap shows the corresponding continuous wavelet transform spectrogram (<40 Hz). An extended frequency spectrogram is shown in Fig. 1-2. Beta bursts are indicated by horizontal lines. Dotted vertical lines mark 1 s intervals.C. PFC PSD (mean±sem) for time intervals when the animal is running on the maze (On Maze) (left) or is immobile at the goal (At Goal) (right). The aperiodic component shown in gray. An extended frequency version is shown in Fig. 1-3 D. CA1 power spectral density (mean±sem) for time intervals when the animal is running on the maze (left) or is immobile at the goal (right). The aperiodic component shown in gray. An extended frequency version is shown in Fig. 1-3. The peak marked by * in the CA1 PSD at ∼15 Hz corresponds to the first harmonic of theta (Scheffer-Teixeira & Tort, 2016). E. PFC PSDs with aperiodic component subtracted. F. CA1 PSDs with aperiodic component subtracted. The beta frequency range is shown in the inset. G. PFC difference between goal and maze PSDs (orange) with 95% confidence interval shown in gray. The beta frequency range is shown in the inset. H. CA1 difference between goal and maze PSDs (blue) with 95% confidence interval shown in gray. I. Frequency profile for beta bursts grouped by peak power frequency for PFC (orange) and CA1 (blue). Top: the vertical histogram of the mean frequency profiles for bursts at each peak power frequency. Bottom: histograms with the count distribution for bursts within 15-30 Hz, grouped by peak power frequency. Shading corresponds to the probability value at each frequency. Bursts occurring within 5 s after reward location entry are included. J. Cumulative distribution of peak burst frequencies for PFC (orange) and CA1 (blue). Kolmogorov-Smirnov test, p=2.9×10-103. K. Distribution of burst times relative to goal location entry grouped by peak frequency for PFC (orange) and CA1 (blue). L. Cumulative distribution of burst occurrence relative to goal location entry for PFC (orange) and CA1 (blue). Kolmogorov-Smirnov test, p=1.1×10-4.

Although the PSD in CA1 did not show a peak in the beta frequency range like the PFC, the continuous wavelet transform from single trials showed that CA1 beta power increased in bursts (Fig. 1B), which is consistent with prior observations of beta bursts in other neocortical regions (Sherman et al., 2016; Shin et al., 2017). To characterize the properties of beta-frequency bursts in PFC and CA1, we extracted burst times from the continuous wavelet transform (Fig. 1-5, Methods). We found the spectral properties of PFC and CA1 bursts differed significantly. PFC bursts were tightly distributed in frequency (median peak frequency: 19 Hz), whereas the peak frequencies of CA1 bursts were distributed across a wider frequency range (median peak frequency: 22 Hz) (Fig. 1I-J, Fig. 1-3 and Fig. 1-4). This explained the peak in the PFC PSD and the lack of such a peak in the CA1 PSD. The burst amplitude profiles also differed between the two regions. In PFC, lower-frequency bursts tended to be higher in amplitude, whereas CA1 bursts were more uniform in size (Fig. 1-6A). Burst durations were similar across the two regions (Fig. 1-6B).

The timing of bursts relative to goal entry also differed between PFC and CA1 (Fig. 1L-M). Bursts tended to start approximately 400 ms after goal location entry, but CA1 bursts were more concentrated within the first two seconds compared with PFC bursts, which showed a sustained pattern. The heatmap shows a clear difference in the burst frequency distribution between the two regions and their distribution over time. Although beta is generally defined as oscillations in the 15-30 Hz range, it has been separated into a lower-frequency beta1 (∼15 Hz) and a higher-frequency beta2 (20-30 Hz), based on work in cortical regions (Baker et al., 1997; MacKay & Mendonça, 1995; Roopun et al., 2008; Roopun et al., 2006). However, the frequency of PFC beta bursts described here is higher than beta1 and lower than beta2. The frequency of CA1 beta bursts identified here (∼22 Hz) is consistent with beta2 previously described in the hippocampus (Berke et al., 2008; Igarashi et al., 2014; Iwasaki et al., 2021). These results indicate that PFC and CA1 beta bursts at the goal have distinct spectral and temporal properties.

Beta oscillations are hypothesized to coordinate neural activity across brain regions (Spitzer & Haegens, 2017) to support cognitive processes including sensory-guided decisions and reinstatement of context-dependent memory patterns (Igarashi et al., 2014; Jayachandran et al., 2023; Kay & Freeman, 1998; Symanski et al., 2022). We therefore asked whether PFC and CA1 beta oscillations are coordinated at the goal. If so, we would expect coherence in the beta-frequency range between PFC and CA1, and similar burst timing in both regions. Visual inspection of LFP signals from PFC and CA1 tetrodes showed that some beta bursts were coincident, while others appeared independent in time (Fig. 2A). To quantify beta-frequency coordination between regions, we first calculated the LFP coherence between tetrodes within each region as well as across regions, for periods when the animal was at the goal. We found higher coherence within PFC or CA1 compared to coherence between PFC and CA1 (Fig. 2B-C, Table 2-1). The PFC coherence within the beta range showed a peak centered at approximately 20 Hz, whereas the CA1 coherence was similar across the entire beta-frequency range. The across-region coherence was significantly weaker but still showed a small peak at ∼20 Hz. Interestingly, coherence was stronger within each PFC hemisphere than across hemispheres, indicating beta dynamics are more coordinated locally within each hemisphere (Fig. 2-1 A-D). There was similar coherence for ipsi-or contralateral PFC and CA1 comparisons (Fig. 2-1 E-G). We then analyzed burst-wise timing and found beta bursts were more coincident within each region compared with across regions (Fig. 2D-E). The cross-correlation of burst peak times showed higher values around 0 s lag within each brain region compared with the cross-correlation between PFC and CA1. Comparing across hemispheres, PFC burst timing was more coincident within each hemisphere than across (Fig. 2-1 H-K) but there was weak coincidence with CA1 irrespective of laterality (Fig. 2-1 L-N).

Beta bursts occur independently in both regions.

A. Beta frequency bursts (shaded) detected across multiple electrodes in PFC (orange) and CA1 (blue) relative to reward location entry (black vertical line). The speed profile is shown in black. Three example trials are shown. B. Coherence (mean±SEM) within PFC (orange), between PFC and CA1 (green), or within CA1 (blue) for periods at the goal location. C. Mean coherence (15-30 Hz) across the groups. Pairwise rank sum test with Benjamini-Hochberg false discovery rate correction: p<0.001 for all pairs. D. Cross-correlation (mean±SD) of burst peak times for bursts detected on tetrodes within PFC (orange), across PFC-CA1 (green) and within CA1 (blue). E. Proportion of the cross-correlation density within ±50 ms lag for PFC (orange), across PFC-CA1 (green) and within CA1 (blue). Pairwise rank sum test with Benjamini-Hochberg false discovery rate correction: PFC versus PFC-CA1 (p=0.003**), PFC versus CA1 (p=0.24n.s.) and PFC-CA1 versus CA1 (p=0.003**). F. Frequency distribution of bursts by peak frequency for independent and coincident bursts in each region. G. Burst peak frequency distribution for coincident and independent bursts. Cumulative distribution burst occurrence relative to goal location entry. Kolmogorov-Smirnov test: PFC (p=0.96n.s.) and CA1 (p=0.002**).

Our results so far suggest beta-frequency oscillations are distinct across PFC and CA1 based on their frequency distributions and low burst coincidence. Despite these differences, it is possible that a subset of coincident PFC and CA1 bursts could coordinate activity across the two regions. This hypothesis predicts that coincident bursts have similar frequency distributions and are more coherent between the two regions. We therefore classified each burst as “coincident” across regions, if it overlapped in time with a burst in the other brain region. Bursts that did not overlap were classified as “independent” across the regions. Based on this method, we asked whether coincident bursts were more closely matched in their frequency distribution across the regions compared with independent bursts. However, this was not the case, as coincident bursts in CA1 were higher in peak frequency than coincident bursts in PFC (Fig. 2F-G). The difference in frequency distribution for the coincident bursts provides evidence against a common beta-frequency range coordinating PFC and CA1, since there is a significant frequency mismatch. We confirmed our classification of independent and coincident bursts using burst-aligned power and coherence analysis (Fig. 2-2). Independent bursts in one region were not accompanied by significant increases in beta power in the other region (Fig. 2-2A-B). Coincident bursts, by definition, showed the expected increase in power in both regions. We found a similar pattern for coherence (Fig. 2-2D-E). Coherence within each region increased during independent bursts but not in the other region. Coherence in both regions increased during coincident bursts. Coherence across regions also increased for coincident bursts more than independent bursts (Fig. 2-2G-H), however, the coherence value across brain regions was lower in comparison with coherence within each region (Fig. 2-2E vs. Fig. 2-2H).

Having established the properties of beta oscillations in PFC and CA1, we next asked how goal-related hippocampal-neocortical beta oscillations changed across task learning. We therefore determined the relationship between beta properties in PFC and CA1 with either days of training or performance in each task. Animals performed better on the Y alternation task than the F alternation task over 5 days (Fig. 3A). By day 5, animals showed clear evidence of learning the Y maze task, with performance reaching greater than 80%. Animals struggled to learn the F maze, with performance remaining close to 50%, which is at chance for the alternation task, making the F maze task a non-learning control condition. Although the performance on the F maze task showed a statistically significant increase across days, the low marginal R2 indicates that the change across days was weak. This performance difference allowed us to compare beta dynamics between tasks with distinct learning outcomes under comparable levels of familiarity, since the animals were trained on both tasks over the same 5-day period. In both tasks, PFC beta-burst power increased over the five days, whereas CA1 beta-burst power decreased (Fig. 3B). We repeated the same set of analyses by partitioning the data based on performance quintiles (Fig. 3D), which allowed us to directly relate beta activity properties with performance. PFC beta-burst power increased with improved performance (Fig. 3D). CA1 beta burst power decreased with improvements in performance (Fig. 3D). These changes were observed on the Y maze task, which animals learned, but not on the F maze, which the animals failed to learn (Fig. 3D). These effects were not due to changes in the animal’s movement over days since we ensured the animal was immobile at the time of the bursts (Fig. 3-1). Thus, we found complementary changes in beta dynamics in PFC and CA1 with familiarity and performance.

Distinct changes in PFC and CA1 burst properties across days of learning.

A. Performance in each maze task by day. Linear model with fixed (day) and random (animal) effects. The marginal R2 (R2m) and conditional R2 (R2c) are shown. B. Burst power in each maze task by day. Linear model with fixed (day) and random (animal) effects. C. Performance in each maze task by performance quintile. Linear model with fixed (performance quintile) and random (animal) effects. D. Burst power in each maze task by performance quintile. Linear model with fixed (performance quintile) and random (animal) effects.

In addition to changes in the strength of beta bursts across days and with improved performance, we asked whether beta coordination between PFC and CA1 changed as a function of day or performance. We found weak effects, indicating PFC and CA1 beta coherence (Fig. 3-2 B and E) and burst coincidence (Fig. 3-3 B and E) did not change as a function of day or performance. Although some of the linear mixed effects models produced statistically significant results, the effect of day or performance on coherence or burst coincidence were weak given the marginal R2 was very low compared with the conditional R2 (Fig. 3-2 B, F and J). The most robust effect was in burst coincidence within CA1, which decreased with performance improvement in the Y maze task, which was learned (Fig. 3-3 C and I).

We next identified temporal changes in beta-frequency-modulated processes in PFC and CA1 with learning. Over days of training, the timing of beta bursts relative to goal entry on unrewarded trials remained unchanged in both PFC and CA1 and on both mazes (Fig. 4A). However, bursts on rewarded trials shifted closer to the time of goal entry for PFC on the Y maze and CA1 on both mazes (Fig. 4A-B). Further, burst timing differed between unrewarded and rewarded trials, with bursts on rewarded trials occurring later than those on unrewarded trials relative to goal entry (Fig. 4A-B, third columns). Similar trends were observed when burst timing was correlated with performance (Fig. 4C-D). We ensured these trends were not due to differences in the movement of the animal at the goal for rewarded or unrewarded trials (Fig. 4-1). The timing effect was comparable in Y and F mazes (Fig. 4-2). Interestingly, the PFC burst peak frequency was lower on rewarded compared with unrewarded trials, which was not the case for CA1 bursts (Fig. 4-3). Across the 5 days, the burst peak frequency also became more consistent, but only when performance improved. The variance in PFC and CA1 burst frequency decreased over days of training for the Y but not the F maze (Fig. 4-4A and C). We ensured this result was not due to differences in burst event counts across days (Fig. 4-4B and D). These results demonstrate that beta frequency modulation in PFC and CA1 has distinct correlates with performance across the five days of training, suggesting changes to beta frequency activity reflect learning-related dynamics in PFC and CA1 networks.

Bursts occur sooner after goal entry over days of training.

A. PFC burst delay distribution relative to goal entry versus days for unrewarded (left column) or rewarded (center column) trials, for the two maze tasks (top or bottom row), and their comparison (right column). B. CA1 burst delay distribution relative to goal entry versus days. C. PFC burst delay distribution relative to goal entry versus performance. D. CA1 burst delay distribution relative to goal entry versus performance.

We next asked how PFC and CA1 spiking activity is modulated at the beta frequency range. We found example PFC and CA1 cells (Fig. 5A-B, Tables 5-1, 5-2 and 5-3) with spike phase-locking preference for the beta frequency range (Fig. 5C-D and Table 5-4). We show example phase-aligned spike rasters to illustrate each cell’s phase locking at each goal location (Fig. 5E-F), as well the corresponding spiking pattern relative to goal entry (Fig. 5G-H). Both PFC (22%) and CA1 (33%) populations showed significant phase locking to beta within each region (Fig. 5I-N). This finding was based on two analysis methods, the Rayleigh test and the von Mises fit (see Methods: Spiking phase locking analysis). The PFC and CA1 cell populations showed phase preferences to the trough of local beta, with PFC cells biased toward spiking during the rising phases, whereas CA1 cells were slightly biased toward the falling phases (Fig. 5K and N). These properties were highly similar between the two maze tasks (Fig. 5-1). Further, we found cells in each region showed higher phase locking to local beta compared with remote beta (Fig. 5O-R). This indicates that spiking in each region is modulated preferentially by local beta oscillations rather than remote beta oscillations. These results provide further evidence supporting that the PFC and CA1 are independently modulated at the beta frequency during the goal period.

PFC and CA1 cells show stronger spiking modulation to local beta.

A. Spike waveforms for example PFC cells. B. Spike waveforms for example CA1 cells. C. Spike triggered power spectrum for example PFC cells. D. Spike triggered power spectrum for example CA1 cells. E. Spike beta phase-locking raster and histogram for example PFC cells. Spike raster is arranged by trial and goal location (color shading). The corresponding mean histogram for each goal location is shown below. The phase of beta is shown in gray. F. Spike beta phase-locking raster and histogram for example CA1 cells. G. Goal location-aligned spiking raster for the example PFC cells. The mean firing rate for each goal location is shown below the spike raster. The corresponding speed at each goal is in the lower row. H. Goal location-aligned spiking raster for the example CA1 cells. I. Distribution of Rayleigh z for PFC. J. Distribution of k from a von Mises distribution fit for significantly beta-modulated cells (Rayleigh p <0.05 in panel b) in PFC. K. Distribution of phase preference for significantly beta-modulated cells in PFC. Red dotted line shows the mean population phase preference. The gray line indicates the phase of beta. L. Distribution of Rayleigh z for CA1. M. Distribution of k from a von Mises distribution fit for significantly beta-modulated cells (Rayleigh p <0.05 in panel b) in CA1. N. Distribution of phase preference for significantly beta-modulated cells in CA1 O. PFC beta phase locking strength using the Rayleigh test (Rayleigh z) to PFC or CA1 beta. Wilcoxon signed-rank test: p=5.4×10-50. P. PFC beta phase locking strength using the von Mises fit (kappa) to PFC or CA1 beta. Wilcoxon signed-rank test: p=5.6×10-43. Q. CA1 beta phase locking strength using the Rayleigh test (Rayleigh z) to CA1 or PFC beta. Wilcoxon signed-rank test: p=5.3×10-55. R. Beta phase locking strength using the von Mises fit (kappa) to CA1 or PFC beta. Wilcoxon signed-rank test: p=1.5×10-45 S. Goal location classification based on mean phase preference at each goal (schematic on the left). Bar plot shows the proportion of the PFC population (non-beta or beta-modulated) that distinguished goal identity based on mean phase preference at each goal. Fisher’s exact test: p=0.19. T. Goal location classification based on mean firing rate each goal (schematic on the left). Bar plot shows the proportion of the PFC population (non-beta or beta-modulated) that distinguished goal identity based on mean firing rate at each goal. Fisher’s exact test: p=0.042. U. Bar plot shows the proportion of the CA1 population (non-beta or beta-modulated) that distinguished goal identity based on mean phase preference at each goal. Fisher’s exact test: p=0.70. V. Bar plot shows the proportion of the CA1 population (non-beta or beta-modulated) that distinguished goal identity based on mean firing rate at each goal. Fisher’s exact test: p=0.66. W. Mean spiking speed for PFC cells based on beta modulation. Wilcoxon rank sum test: p=0.41. X. Spiking distance to nearest goal for PFC cells based on beta modulation. Wilcoxon rank sum test: p=0.0006. Y. Spatial information for PFC cells based on beta modulation. Wilcoxon rank sum test: p=0.18. Z. Mean spiking speed for CA1 cells based on beta modulation. Wilcoxon rank sum test: p=0.29 AA. Spiking distance to nearest goal for CA1 cells based on beta modulation. Wilcoxon rank sum test: p=0.13 BB. Spatial information for CA1 cells based on beta modulation. Wilcoxon rank sum test: p=0.013.

We next determined whether beta-phase-locked or non-phase-locked cells had distinct task correlates. Because we found beta oscillations occurred at goals, we first asked whether phase locking preference or firing rate distinguished between the goal locations, which could suggest phase or rate coding of goal identity. Using a Gaussian Naïve Bayes model to classify goal identity based on the beta phase for spikes, we found the phase of PFC spiking was a poor differentiator of goal identity (Fig. 5S). PFC firing rate differed between goals, with a small but significantly greater proportion of beta-phase-locked cells with this property compared with non-phase-locked cells (Fig. 5T). For CA1, a larger proportion of the cell population differentiated the goals based on spiking phase (Fig. 5U). We note that for CA1, even cells that were classified as not modulated by beta showed differences in phase locking between goals. This is explained by our beta phase modulation classification being based on spiking from all goals rather than individual goals, where a non-beta-modulated cell could show spike phase preference at one of the goals. Similarly, for firing rate, a large proportion of CA1 cells had different firing rates at each goal, and the proportions were similar between phase locked and non-phase-locked cells (Fig. 5V). In additional analyses of task firing correlates, we found PFC beta-modulated cells fired closer to goals compared with non-phase-locked cells (Fig. 5X), an effect driven by the F maze (Fig. 5-2B). PFC and CA1 beta phase-locked cells had more spatial information compared with non-phase-locked cells on the Y maze (Fig. 5-2C). Beta phase locking status did not identify differences in the speed preference of the cells (Fig. 5W and Z, and Fig. 5-2A).

We next asked how changes in PFC and CA1 beta oscillations are related to another prominent neural process supporting learning: awake hippocampal SWRs. Hippocampal SWRs also occur at goal locations (Ambrose et al., 2016; Carey et al., 2019; Diba & Buzsaki, 2007; Foster & Wilson, 2006; Kleinman & Foster, 2025; Singer & Frank, 2009) and are hypothesized to coordinate distributed representations (Jadhav et al., 2016; Shin et al., 2019; Tang et al., 2021; Yu et al., 2017; Yu et al., 2018) to support the formation and use of memory during spatial learning (Cheng & Frank, 2008; Deceuninck & Kloosterman, 2024; Dupret et al., 2010; Gupta et al., 2010; Jadhav et al., 2012; Pfeiffer & Foster, 2013; Singer et al., 2013; Singer & Frank, 2009). Since both beta and SWRs occur at goal locations, we asked how beta-frequency dynamics within PFC and CA1 networks are related to hippocampal SWR dynamics. Consistent with previous findings, we found SWRs tended to occur several seconds after the animal reached the goal (Fig. 6A-B) (Ambrose et al., 2016; Carey et al., 2019; Diba & Buzsaki, 2007; Foster & Wilson, 2006; Kleinman & Foster, 2025; Singer & Frank, 2009). SWR power (Fig. 6-1A) and rate (Fig. 6-1B), but not duration (Fig. 6-1C) decreased across days, consistent with prior findings (Cheng & Frank, 2008). We ensured the animals were immobile (Fig. 6-1D), and the goal visit count (Fig. 6-1E) and duration (Fig. 6-1F) were comparable over days.

Goal location beta bursts and SWRs are coordinated at multiple timescales.

A. PFC (orange) and CA1 (blue) LFP relative to goal entry (black vertical line). Beta frequency filtered signal is shown below the LFP. Ripple frequency (150-250 Hz) filtered CA1 signal is shown in black. The corresponding speed of the animal is shown in the bottom row. B. Goal-entry-aligned histogram for CA1 SWRs. Median and interquartile range are shown. C. Goal-entry-aligned histogram for PFC beta bursts. D. Goal-entry-aligned histogram for CA1 beta bursts. E. Cross-correlation between PFC beta bursts and CA1 SWRs (±5 s, 1 s bins). Wilcoxon signed-rank test p=5.2×10-8. F. Cross-correlation between CA1 beta bursts and CA1 SWRs (±5 s, 1 s bins). Wilcoxon signed-rank test p=7.1×10-8. G. Cross-correlation between PFC beta bursts and CA1 SWRs (±500 ms, 100 ms bins). Bins with significant deviations (with Benjamini-Hochberg false discovery rate correction) from the mean are indicated with *. H. Cross-correlation between CA1 beta bursts and CA1 SWRs (±500 ms, 100 ms bins). Bins with significant deviations (with Benjamini-Hochberg false discovery rate correction) from the mean are indicated with *. I. SWR-aligned PFC spectrogram (100-300 Hz) and corresponding ripple band power (150-250 Hz). J. SWR-aligned CA1 spectrogram (100-300 Hz) and corresponding ripple band power (150-250 Hz). K. SWR-aligned PFC spectrogram (10-30 Hz) and corresponding beta-band power (15-30 Hz). L. SWR-aligned CA1 spectrogram (10-30 Hz) and corresponding beta-band power (15-30 Hz).

We then found distinct temporal coordination patterns between beta bursts and SWRs. In contrast to the delayed onset of hippocampal SWRs, beta bursts in PFC and CA1 occurred closer to the time of goal entry (Fig. 6C-D); thus, on the timescale of seconds, beta bursts in each region occurred earlier than SWRs. We confirmed this relationship occurred on a single-trial timescale by computing the cross-correlation between the timing of beta bursts in each region and SWRs (Fig. 6E-F). We next determined whether beta bursts and SWRs were temporally coordinated on a finer timescale. On a shorter timescale of tens of milliseconds, we found an inverse coordination pattern between beta bursts and hippocampal SWRs in each brain region (Fig. 6G-H). PFC beta bursts were less frequent at the time of hippocampal SWRs (Fig. 6G), whereas CA1 beta bursts and SWRs were more coincident (Fig. 6H). This pattern of coordination was further supported by SWR-aligned spectrograms (Fig. 6I-L). PFC beta-band power showed a decrease at the time of hippocampal SWRs (Fig. 6K) whereas CA1 beta-band power showed an increase (Fig. 6L). These patterns remained consistent when the Y and F maze data were analyzed separately (Fig. 6-2). It is possible that the CA1 beta band power around the time of SWRs results from accompanying sharp waves (Buzsaki et al., 1983); however, our burst timing analysis showed that the CA1 beta bursts were more frequent before the onset of SWRs (Fig. 6H). We also found SWR band power showed phase coupling to CA1 beta but not PFC beta (Fig. 6-3). In CA1, SWR power was maximum at the peak of local beta (Fig. 6-3 C-D). Thus, the temporal coordination between beta bursts and SWRs occurs on the second and millisecond timescales and supports both markers being complementary neural signatures for learning-related processes.

Since our results suggest that goal location beta frequency activity and SWRs correspond to temporally coordinated processes, do they modulate distinct populations of PFC and CA1 cells? In PFC and CA1, we found cells modulated by both SWRs and beta (during the 5 s after goal entry) (Fig. 7A, B, G and H). In PFC, ∼22% of cells showed modulation to beta and ∼30% showed modulation to SWRs. The proportion of PFC cells that are modulated by both SWR and beta (8%) is not significantly different from the expected proportion (6.6%) under the assumption that the population is independently modulated by SWR and beta (30%ξ22%) (Fisher’s exact test, p=0.09). We found PFC cells modulated by both beta and SWRs (SWR+β+) had distinct firing properties. The SWR+β+ population had a positive relationship between the spike-beta phase preference and the direction of SWR modulation (Fig. 7C). PFC cells with spiking preference for the falling phases of the beta cycle tended to show spiking inhibition during SWRs (negative SWR modulation index) (Fig. 7A, lower row) whereas cells with spiking preference for the rising phases of beta tended to be excited during SWRs (positive SWR modulation index) (Fig. 7A, upper row). As a control, this relationship was not observed for PFC cells that were not modulated by beta but modulated by SWRs (SWR+β-) (Fig. 7C, lower panel, Fig. 7-1A). The SWR+β+ PFC cell population was further distinguished from the SWR+β- population by its strong positive relationship with goal location firing. Negatively SWR-modulated cells were active closer to goals whereas positively SWR-modulated cells were active further away from goals (Fig. 7D, upper panel). This relationship was absent for SWR+β- cells (Fig. 7D, lower panel). SWR-modulated PFC cells showed a positive relationship with speed, where negatively modulated cells spiked when the animal was at a lower speed and positively modulated cells spiked when the animals were moving faster (Fig. 7E), which is consistent with prior work (Jadhav et al., 2016; Yu et al., 2017). This relationship did not depend on beta modulation status. Similarly, SWR-modulated PFC cells had comparable spatial information irrespective of beta modulation status (Fig. 7F). Additional task correlate analysis showed that beta modulation did not further segregate SWR-modulated PFC cells based on phase or firing rate coding of goal locations (Fig. 7-2A and B).

SWR and beta modulation in PFC and CA1.

A. Four example PFC cells showing SWR-aligned spiking raster and histogram, and spike beta-phase-locking histogram. Top two cells show spiking excitation around SWRs, and bottom two cells show spiking inhibition around SWRs. The SWR modulation index (MI) is displayed. In the spike beta-phase-locking histogram, the distribution is duplicated to improve visualization. The peaks and troughs for the beta cycle are shown in white and the phase locking strength (k) is displayed. B. PFC beta modulation strength (kappa) versus SWR modulation index. The four combinations of SWR (S) and beta (β) modulation (+: modulated and -: not modulated) and their corresponding proportions are indicated. C. PFC SWR modulation index versus beta phase locking preference for beta and SWR-modulated cells (β+S+) (upper) and beta-modulated but not SWR-modulated cells (β-S+) (lower). The regression line is shown in cyan. The gray shading indicates the peak and trough of beta. Two cycles are repeated for clarity. D. Spiking distance to nearest goal versus SWR modulation index. E. Spiking speed versus SWR modulation index. F. Spatial information versus SWR modulation index. G. Four example CA1 cells showing SWR-aligned spiking raster and histogram and spike beta-phase-locking histogram. H. CA1 beta modulation strength (kappa) versus SWR modulation index. The four combinations of SWR and beta (β) modulation (+: modulated and -: not modulated) and their corresponding proportions are indicated. I. CA1 SWR modulation index versus beta phase locking preference for beta and SWR-modulated cells (β+S+) (upper) and SWR-modulated but not beta-modulated cells (β-S+) (lower). The regression line is shown in red. The gray shading indicates the peak and trough of beta. Two cycles are repeated for clarity. J. Spiking distance to nearest goal versus SWR modulation index. K. Spiking speed versus SWR modulation index. L. Spatial information versus SWR modulation index.

For CA1, we found modulation patterns distinct from PFC. We repeated these analyses for the CA1 population and found the opposite relationship between beta phase preference and SWR modulation compared with PFC cells (Fig. 7I). In CA1, ∼33% of cells showed beta phase modulation and ∼56% of cells showed modulation to SWRs. The subset of CA1 cells that are modulated by both SWR and beta (SWR+β+, 18%) was similar to the expected proportion (19%) under the assumption of independence (33%ξ56%) (Fisher’s exact test, p=0.62). For this population, cells with spiking preference for the falling phases of the beta cycle tended to have a positive SWR modulation index, indicating stronger excitation during SWRs (Fig. 7I). As a control, this relationship was not observed for the SWR+β- CA1 cells (Fig. 7I, lower panel, Fig. 7-1B). Unlike PFC cells, SWR-modulated CA1 cells showed a strong positive correlation between the direction of SWR modulation and spiking distance relative to the goal and speed, irrespective of beta modulation patterns (Fig. 7J-K). This is consistent with known properties of dorsal CA1 cells, where cells with place fields on the maze leading to goals are reactivated and immobility cells active at goals are suppressed (Jin et al., 2024; Kay et al., 2016; Yu et al., 2017). SWR+β-CA1 cells showed a negative correlation with spatial information, however, the relationship is weak (Fig. 7L, lower panel). We also determined whether SWR and beta modulation status identified cells that could distinguish goal identity, either based on phase locking preference or firing rate at goals. However, these were similar irrespective of beta or SWR modulation status (Fig. 7-2A-B). Similar to PFC cells, beta modulation did not further segregate the SWR-modulated population for phase or firing rate coding of goal locations (Fig. 7-2A and B).

Our results show that beta oscillations modulate a subpopulation of PFC neurons active at the goal, and some of these cells are coordinated with hippocampal SWRs. The PFC subset may not be coordinated with the hippocampus at the beta frequency, but they can still be linked with relevant hippocampal representations during SWRs. This subset of PFC cells expressed a more faithful recapitulation of the task structure relative to goals than the rest of the PFC population, which may represent a functional subpopulation engaged in representing the ongoing task structure.

Discussion

Our results show beta-frequency oscillations occurred after entry to goal locations in both CA1 and PFC. However, several properties suggest that the two brain regions are weakly coordinated at the beta frequency. First, we found that beta bursts had distinct spectral and temporal properties in each brain region, where PFC beta bursts were lower in frequency than CA1 bursts. Second, PFC burst amplitude increased with performance, whereas CA1 burst amplitude decreased. This was observed when animals learned the task (Y maze) but not when they failed to learn the task (F maze). The pattern of learning-related changes for PFC bursts was also opposite to those of SWRs. Lastly, beta activity and SWR-modulated PFC cells identified a specific subpopulation with distinct task correlates, but not in CA1.

Our results suggest that the neocortex and hippocampus may participate in distinct functional networks at the goal, and beta oscillations may support each region’s interactions with its respective network partners at distinct frequencies within the beta band. This contrasts with other task periods where the PFC and hippocampus are part of the same functional network modulated by beta or other frequencies. For example, during spatial navigation, PFC cells show strong phase locking to hippocampal theta (Hyman et al., 2005; Jones & Wilson, 2005; Siapas et al., 2005; Yu & Frank, 2021; Zielinski et al., 2019), a frequency band that has been hypothesized to reflect coordination of distinct task-related representations across these structures during the execution of cognitive tasks. During olfactory sensory decision-making, beta-frequency coherence is observed across neocortical and hippocampal networks, which is thought to coordinate associative memory representations (Fourcaud-Trocmé et al., 2019; Igarashi et al., 2014; Jayachandran et al., 2023; Kay & Freeman, 1998; Martin et al., 2007; Sheriff et al., 2021; Symanski et al., 2022). During awake quiescence, activity in neocortical and hippocampal networks is transiently synchronized during SWRs (Jadhav et al., 2016; Remondes & Wilson, 2015; Shin et al., 2019; Tang et al., 2017; Yu et al., 2017; Yu et al., 2018). Coordination at these frequencies and task periods demonstrates hippocampal-neocortical interactions play a central role in supporting diverse cognitive functions. Our results show that the neocortex and hippocampus may be weakly coupled during the goal period at the beta frequency, and each region may be engaged with its own set of functionally connected networks (Cohen et al., 2011). This was surprising given the extensive coordination across these brain structures during other important task periods. We had expected that stronger coordination during this period would be important for learning-related changes across neocortical-hippocampal networks, similar to SWRs that occur at the goal.

Our results point to another insight on the complementary changes in hippocampal-neocortical beta dynamics, on multiple timescales. First, we found an inverse pattern in the hippocampus and PFC during learning, which spans the timescale of days. These results indicate that beta-frequency oscillations in the neocortex and hippocampus are markers of complementary learning-induced changes in these networks, pointing to the contrasting contributions of the hippocampus and neocortex to learning. The hippocampus is hypothesized to play an important role in rapid changes during early learning, whereas the neocortex gradually changes on a longer timescale (Buzsáki, 1996; Mcclelland et al., 1995). In the hippocampus, SWRs are a neural signature of early learning, where larger amplitude events are observed during novel experience (Cheng & Frank, 2008). With learning, these events become less prominent. We found similar patterns for goal-related beta activity in the hippocampus, which decreases with performance. This mirrors novelty-related beta activity in the hippocampus that occurs during exploration of new environments or objects, which decreases with familiarity (Berke et al., 2008; França et al., 2014; França et al., 2021; Rangel et al., 2015). In contrast to hippocampal beta oscillations, we hypothesize that beta activity in the PFC reflects the reconfigured state for a given task as learning progresses, where beta amplitude and frequency uniformity increases with experience and performance. This increase parallels findings in the olfactory bulb and piriform cortex, where the strength of beta during odor sampling increases with performance proficiency in olfactory sensory decision-making tasks (Cohen et al., 2015; Fourcaud-Trocmé et al., 2019; Gervais et al., 2007; Martin et al., 2007; Martin et al., 2004), the ventral striatum (Howe et al., 2011), as well as the emergence of beta oscillations in the basolateral amygdala with the development of reward preference (Amaya et al., 2024).

On the timescale of seconds, we found beta bursts typically precede hippocampal SWRs after goal entry (Fig. 6A-F). Beta bursts in PFC and CA1 occurred within a second after goal entry and became less frequent. In contrast, hippocampal SWRs became more frequent over several seconds after goal entry. The temporal segregation likely points to distinct goal-related processes. Immediately after goal entry, PFC and CA1 are separately engaged with their own beta-modulated networks, which then transitions to strong coordination between the two regions during hippocampal SWRs to support coordinated hippocampal-neocortical reactivation. Interestingly, on the timescale of milliseconds, ongoing PFC beta oscillations are reduced during hippocampal SWRs (Fig. 6K), whereas CA1 beta power increases around the time of SWRs (Fig. 6L). This is again consistent with the hypothesis that PFC and CA1 beta are signatures for functionally distinct and uncoordinated processes between the two regions. When hippocampal SWRs occur, a switch in the hippocampal-neocortical state occurs, reducing ongoing neocortical beta oscillations in favor of reactivation-related coordination across hippocampal-neocortical networks (Jadhav et al., 2016; Yu et al., 2017). Despite our identification of goal period timing relationships between beta oscillations and hippocampal SWRs, further investigation is needed to determine the functional contribution of beta modulation during this period.

We found that beta modulation status identifies a specific subpopulation of SWR-responsive PFC cells with distinct task firing correlates. Prior work shows that the PFC firing during awake hippocampal SWRs can reactivate movement representations (Shin et al., 2019), where SWR-excited PFC cells prefer to spike during movement and SWR-inhibited PFC cells prefer to spike during immobility (Jadhav et al., 2016; Yu et al., 2017). Our results are consistent with these findings (Fig. 7E). Interestingly, beta and SWR-modulated PFC cells had a distinguishing feature, where they better recapitulate the distance to goal relationship than non-beta-modulated PFC cells (Fig. 7D). This suggests PFC cells with beta modulation may represent a cell subpopulation participating in the expression of goal-related states, beyond movement coding alone.

The difference in the frequency distribution of hippocampal and PFC beta raises the question of whether cortical and hippocampal beta have similar origins. Given that beta oscillations have been found in connected cortical, subcortical, and hippocampal regions, do they arise from similar mechanisms? Experimental evidence and computational models indicate that beta oscillations could be generated locally within a network (Bitzenhofer et al., 2017; David et al., 2015; Fourcaud-Trocmé et al., 2011; Osinski & Kay, 2016; Roopun et al., 2008; Roopun et al., 2006; Sherman et al., 2016; Womelsdorf et al., 2014). Beta was originally partitioned into a lower-frequency ∼15 Hz (beta1) band and a higher-frequency 20-30 Hz (beta2) band (Baker et al., 1997; MacKay & Mendonça, 1995; Roopun et al., 2008; Roopun et al., 2006). Our observations in CA1 are consistent with beta2 (23-30 Hz) (Berke et al., 2008; Igarashi et al., 2014; Iwasaki et al., 2021), with our observed beta bursts having a median peak frequency of approximately 22 Hz. The PFC beta we observed centers around 19 Hz with a narrow frequency distribution. This is closer to beta1 but is higher than the ∼15 Hz beta1 reported in various neocortical regions. Could the distinct beta frequency distribution in these regions point to distinct origins? PFC beta may be generated locally or from a brain region with a lower frequency beta generator. The wider distribution of CA1 beta burst peak frequencies suggests hippocampal beta may arise from multiple local or remote inputs from a variety of brain regions, each with a range of central frequencies above 20 Hz. These oscillations may be relayed to CA1, possibly via the midline thalamic regions, including the nucleus reuniens (Jayachandran et al., 2023), or via the entorhinal cortex (Kay & Freeman, 1998; Kay et al., 1996).

There is increasing evidence for the contribution of beta oscillations to a wide range of cognitive processes (Lundqvist et al., 2024; Miles et al., 2023), expanding from prior work focused on sensorimotor control (Barone & Rossiter, 2021; Engel & Fries, 2010) and sensory perception (Kay, 2014). While beta has been implicated in coordinating neural activity across brain networks, including the hippocampus and neocortex (Igarashi et al., 2014; Jayachandran et al., 2023; Spitzer & Haegens, 2017; Symanski et al., 2022), we identified an important period for learning during which the hippocampus and PFC could be modulated independently by distinct frequency bands within the beta range. Identifying functional subnetworks and when they coordinate can provide further insight into the temporal dynamics of distributed neural processes supporting cognition. An important future direction will be to test the causal contribution of beta frequency dynamics across distributed brain networks in learning.

Methods

Animals

Long-Evans rats (Charles River Laboratories, 6 males, 4-10 months) were used for the study. All procedures were performed under approval by the university’s Institutional Animal Care and Use Committee, according to the guidelines of the Association for Assessment and Accreditation of Laboratory Animal Care.

Neural implants

Animals were chronically implanted with neural recording devices (Voigts et al., 2020) with up to 32 adjustable tetrodes. Up to 16 tetrodes targeted the CA1 region of the left hippocampus (AP −3.8, ML −2.75 relative to bregma), and up to 16 tetrodes targeted both hemispheres of the prelimbic region of rat PFC (AP +3.5, ML ±1.5 relative to bregma). The PFC bundle was angled at 15° towards the midline to avoid the superior sagittal sinus. A ground screw was placed in the skull above the right cerebellum. Tetrodes were gradually adjusted into the target regions over two to three weeks after surgery, based on LFP markers and estimated depth of the tetrodes. Once the tetrodes were within the target brain regions, small adjustments of 25 μm, when necessary, were made to isolate single cells. Tetrode locations were verified at the end of the experiment. To mark the tip of the tetrode, electrolytic lesions were induced for each tetrode (100 μA, 2 s). Nissl staining was used to verify the location of the lesion was within the intended area.

Spatial learning apparatus

All tracks and mazes were constructed from acrylic (black DP-9, 8 cm wide, 4 cm tall walls, 76 cm above floor level). Reward delivery in all tasks was automated. Infrared sensors were placed at the ends of the maze arms to detect the rat’s arrival and to trigger the reward. Reward was delivered by a syringe pump (100 μl at 20 ml/min, NE-500, New Era Pump Systems Inc, New York, USA).

Goal entry is defined as the time when the animal reaches the reward delivery device and breaks the infrared beam. All timestamps for beam breaks and for when the beam is restored at each goal are logged by the recording system.

Behavioral training

Animals were food restricted to >85-90% of their baseline body weight. For pretraining, animals were first trained to forage for reward in a black open field box (H: 31 cm, W: 61 cm, L: 61 cm), for 10 minutes per day for 3 days. The reward was 70% evaporated milk (Carnation) with 5% sucrose. The reward was randomly dropped inside the open box to encourage foraging. The next pretraining phase involved the rats learning to run back and forth on an elevated linear track (60 cm) to consume reward from the ends. Animals were trained until a performance criterion of 20 rewards per session (10-minute sessions, three times daily, 4 to 7 days).

For the tasks on the Y and F shaped mazes, the animals needed to learn a specific sequence of goal visits to be rewarded on every trial. There were three goal locations in each maze, which were wells at the end of the arms. The animal needed to alternate visits between wells 2 and 3 via well 1 (1, 2, 1, 3, 1, 2, 1, 3 …), similar to previously described spatial alternation tasks (Frank et al., 2000; Jadhav et al., 2012; Kim & Frank, 2009; Maharjan et al., 2018). For example, if the rat was at well 1, it would be rewarded next at well 2 only if it had previously visited well 3. Similarly, if the rat was at well 1, it would be rewarded next at well 3 only if it had previously visited well 2. If the rat was at wells 2 or 3, it would be rewarded only by going to well 1. Reward delivery was controlled using custom scripts written for the data recording system.

Data recording

Data were collected using the Trodes data acquisition system (SpikeGadgets LLC, California, USA). Neural data were recorded using a 128-channel digitizing headstage. The head stage was tethered via a commutator to the main data acquisition system. The commutator and tether were supported by a passive suspension system. Raw data were recorded at 30 kHz. Local field potential (<300 Hz) was extracted at 1,500 Hz. Time-synchronized video (1456 x 1088, 30 Hz) was recorded using Allied Vision Manta G-158C cameras.

Behavioral analysis

DeepLabCut (Mathis et al., 2018) was used to track the animal in the video. The spatial coordinate for the neck of the animal was used as the animal’s location and to calculate the animal’s speed. Performance of the animals was derived using a state-space model that estimated the probability of the animal making a correct choice (Smith et al., 2004). We then calculated the mean performance for each day, and also grouped performance by quintiles.

Data and statistical analysis

We used Python packages for signal processing: Elephant electrophysiology toolkit (https://elephant.readthedocs.io), fooof tools (https://fooof-tools.github.io/fooof/), GhostiPy (https://github.com/kemerelab/ghostipy); circular statistics: PyCircStat (https://github.com/circstat/pycircstat); and statistics: SciPy (scipy.org), statsmodels (statsmodels.org) and pingouin (pingouin-stats.org).

We used linear mixed-effects models (statsmodels mixedlm) to perform regression analysis to account for the contribution of fixed effects from the independent variable and the random effects from individual variation between animals. We report the marginal R2 (R2m), which is the proportion of variance explained by the fixed effects and the conditional R2 (R2c), which is the proportion of variance explained by the entire model (fixed and random effects) (Nakagawa & Schielzeth, 2013).

Spectral analysis

We calculated the power spectral density (PSD) using the Welch function, with 1 s segments with 50% overlap (scipy.signal.welch). We calculated coherence using the coherence function with 1 s segments and 50% overlap (scipy.signal.coherence). We isolated the periodic component of the PSD by subtracting the aperiodic component using Fitting Oscillations & One-Over-F (fooof-tools) (Donoghue et al., 2020). This was done for each behavior session, and the mean was calculated across all sessions per day or for sessions on each maze separately. For periods at the goal, we only included 1 s segments when the animal was immobile (speed < 0.5 cm/s). For periods on the maze, we only included 1 s segments when the animal was mobile (speed ≥ 4 cm/s). We calculated the difference in PSD between goal and maze periods by subtracting the periodic component PSDs. A 95% confidence interval for the difference was generated using a permutation test (1000 shuffles) where we permuted the maze or goal identity of each PSD.

We repeated the spectral analysis using continuous wavelet transform (CWT) for a ±3 s window centered on goal entry, using a generalized Morse wavelet (γ = 4, β = 7.5) for 2–100 Hz in 1 Hz steps (GhostiPy cwt). We calculated the mean PSD for the 3 s window on the maze before goal entry and the 3 s after goal entry, as well as the difference between the two. This was done for each behavior session, and the mean was calculated across all sessions per day or for sessions on each maze separately. A 95% confidence interval for the difference was generated using a permutation test (1000 shuffles) where we permuted the behavioral state (maze or goal) of each spectrogram.

Beta burst detection

Local field potential (LFP) analysis was performed using the Elephant electrophysiology toolkit (Denker et al., 2024). We applied CWT to extract time-resolved spectral power for 0.5–130 Hz in 1 Hz steps. This was done using the Ghostipy package (Chu & Kemere, 2021). The average power in the beta-band (15–30 Hz) isolated from the CWT spectrogram was used to detect bursts using a dual-threshold amplitude method (Feingold et al., 2015). Bursts were defined as periods when the beta power envelope exceeded 1 standard deviation (SD) above the mean power with a minimum duration of 100 ms. The onset and offset of each burst were marked by the first upward and downward crossings of 0.5 SD above the mean, respectively. Burst peaks were characterized by the z-scored amplitude at the power maximum. Events separated by less than 100 ms were combined. Periods where the maximum LFP signal amplitude across all tetrodes exceeded 10 standard deviations were classified as noise and excluded from analyses.

We identified the time at which the burst had its peak power, the magnitude of the peak power, and the frequency at the time of peak power. This was done by isolating the CWT spectrogram for each burst and extracting the time and frequency corresponding to the maximum power value within the burst. Bursts were classified as independent or coincident based on overlap. Independent bursts are those without overlap in time, whereas coincident bursts overlap in time.

Burst-aligned power and coherence analysis

Event-aligned power and coherence within and across PFC and CA1 (Fig. 2-2) were computed using the multi-taper method (2 tapers with a time-bandwidth product of 2) (Python implementation of the coherence function from chronux.org) (Mitra & Bokil, 2008). We calculated the coherence in a 1 s window centered at the peak of each burst (500 ms moving window, 5 ms steps). For within-region coherence, the coherence value was the mean for all pairs of tetrodes. To determine the change in coherence around the peak of the burst, a baseline was defined as the mean coherence in the −500 to −375 ms interval relative to the center of the burst. We used linear mixed-effects models (statsmodels.mixedlm) to determine the relationship between burst power or frequency variance and performance (by day or quintiles) while accounting for differences between animals.

Sharp wave ripple detection

Sharp-wave ripples (SWRs) were identified based on previous methods (https://github.com/Eden-Kramer-Lab/ripple_detection) (Kay et al., 2016). LFP from tetrodes in the dorsal CA1 region (oriens and pyramidale) was filtered in the ripple frequency band (150–250 Hz). The median of the smoothed envelope (Gaussian kernel σ = 4 ms) was calculated. We used a threshold of 2 standard deviations for detecting SWRs, with a minimum duration of 50 ms and separation of 50 ms. Only SWR events occurring at speed <4 cm/s and at the goal locations were included. We filtered out detection artifacts due to LFP noise by excluding time intervals in which the LFP signal magnitude exceeded 10 standard deviations across any tetrode.

Spike sorting

We used MountainSort (Chung et al., 2017) with manual curation to isolate single cells. The quality of well-isolated cells was determined by having an interspike interval >2 ms, consistency of the waveforms, and the stability of the peak voltage values over the sessions.

Cell-type classification

We classified each single unit as a pyramidal cell or an interneuron using established methods based on firing rate (Jadhav et al., 2016; Kay et al., 2016), trough-to-peak width (Petersen et al., 2021), and autocorrelation function mean (Kay et al., 2016). The thresholds for classification are shown in Table 5-1.

Spike-aligned spectrogram

For each cell, the spike-aligned spectrogram (Fig. 5C-D) was generated by extracting the CWT at the time of each spike and taking the average.

Spike phase-locking analysis

For each tetrode, the phase of beta oscillations was extracted by first applying a band-pass filter to the raw LFP signal within the beta band range (15-30 Hz, acausal 4th order Butterworth), followed by a Hilbert transform using the Elephant toolkit (Denker et al., 2024). For each cell, we calculated the phase at each spike relative to the beta phase for every available tetrode, which allowed us to determine the cell’s phase-locking preference to beta within the same brain region (local) and the other region (remote). We used two metrics to quantify the strength of spike phase-locking to beta. First, we performed the Rayleigh test for circular non-uniformity and calculated the Rayleigh-z value using PyCircStat (Berens, 2009). Second, we fitted a von Mises distribution to the phase preference histogram to estimate the mean phase and the concentration parameter (kappa) (scipy.stats.vonmises). To determine the local phase-locking strength of a given cell, we chose the maximum values of the phase-locking metric across all tetrodes within the same region as the cell. Similarly, we calculated the remote phase-locking strength as the maximum phase locking value across all tetrodes in the other brain region. Only spikes that occurred when the animal was at the goal location were included for the phase-locking analysis.

Task correlates for spiking activity

To determine if each cell’s beta phase preference could distinguish between goal locations, we used a Gaussian Naïve Bayes model (scikit-learn, GaussianNB). For the model, the input feature was the mean spike beta phase preference (5 s after goal entry), and the label was the goal identity. We determined the statistical significance of the classification accuracy using a permutation test (scikit-learn, permutation_test_score, 1000 shuffles and 3-fold cross-validation) where the goal identity labels were permuted. We used the same method to determine whether a cell’s mean firing rate could distinguish between the goal locations.

The mean spiking speed is the mean speed for all spikes for each cell. The distance to goal is the mean of the shortest linear distance from the location at each spike to a goal, accounting for the path geometry of the maze. The spatial information is the entropy of a cell’s distance-normalized firing pattern on all trajectories. Higher entropy indicates greater non-uniformity in firing patterns across trajectories.

SWR-aligned spectrogram

We applied CWT to extract time-resolved spectral power for 0.5–300 Hz in 1 Hz steps using the same parameters as other CWT analysis described earlier. We then aligned the PFC or CA1 spectrogram relative to the start of CA1 SWRs for each behavior session, and calculated a session mean. The mean spectrogram for all sessions for each region was shown in Fig. 6. We repeated this for all sessions on each maze in Fig. 6-2. From these spectrograms, we quantified the mean power in the SWR band power (150-250 Hz) or beta band (15-30 Hz).

SWR modulation of spiking

SWR modulation was determined based on previously described methods (Jadhav et al., 2016; Rothschild et al., 2017; Yu et al., 2017; Yu et al., 2018). For each cell, a SWR event-aligned spiking histogram was created for a 1 s window centered on the start of each SWR. A modulation index was calculated based on the difference in spiking rate within a 200 ms window after the start of the SWR relative to the baseline within the 1 s window. A circular-shift permutation test was used to determine the significance of the modulation index. A positive modulation index corresponds to an increase in firing around the time of SWRs, whereas a negative modulation index corresponds to a decrease in firing. A permutation test was used to determine the significance.

Phase amplitude coupling

For each session, one tetrode from the pyramidal layer of CA1 and one tetrode from PFC was selected for the analysis. We extracted the power in the SWR frequency band for the time intervals during SWRs, from the CA1 tetrode. We then extracted the corresponding beta phase from the CA1 or the PFC tetrode for those intervals. We then binned the phase values (-π to π, π/8 bins) and assigned the SWR band power to their respective phase bins. For each phase bin, we calculated the mean SWR band power. We then averaged across all sessions on each maze.

To calculate the chance power distribution, we used a permutation test with 5000 shuffles. For each shuffle, we circularly permuted the phase bins for each session and calculated the mean across all sessions. The 99% confidence interval for each phase bin was derived from the permuted means.

Correlation between beta phase locking preference and SWR modulation index or goal-entry firing

We calculated the Pearson’s correlation between a cell’s beta phase preference and its SWR modulation index or goal-entry firing. Goal-entry firing measures the firing latency relative to goal entry and is defined as the time corresponding to 50% of the firing density in the 5 s from goal entry. To estimate the significance of the correlation coefficient, we performed a permutation test. Each cell was classified based on its phase-locking significance to local beta (β+ or β-) and its SWR modulation significance (SWR+ or SWR-). This produces four possible combinations in the population (β-SWR-, β+SWR-, β-SWR+, or β+SWR+). We then permuted the β and SWR significance labels for the population and repeated the calculation. This was done 5000 times to generate a permuted correlation coefficient distribution, which allowed us to estimate the significance to p<0.0002. The p value of the observed correlation coefficient is calculated as the proportion of the permuted distribution that is greater or less than the observed value.